---
title: "Differential expression"
output: html_document
date: "2025-06-19"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/scratch/prj/dn_hamidlab_msc_proj/k24082190/originals")
```

```{r}
#Setting working directory
setwd("/scratch/prj/dn_hamidlab_msc_proj/k24082190/originals")
```

```{r}
#setting custom path
.libPaths("/scratch/prj/dn_hamidlab_msc_proj/Libraries")
```




```{r}
#load the libraries needed
## load dependencies
library(tidyverse)
library(Seurat)
library(SeuratObject)
library(sp)
library(scDblFinder)
library(SingleCellExperiment)
#library(SeuratWrappers)
#library(SeuratDisk)
library(patchwork)
library(clustree)
library(dplyr)
library(XML)
library(Matrix)
library(glmGamPoi)
library(gprofiler2)
library(harmony)
library(purrr)
library(limma)
library(umap)
library(pheatmap)
library(fgsea)
library(clusterProfiler)
library(org.Mm.eg.db)
library(rrvgo)
library(slingshot)
library(destiny)
library(tradeSeq)
```




## loading in the non-OSNs from the wildtype
```{r}
#continue from here
merged_non_OSNs <- readRDS("/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/merged_non_OSNs.rds")
```



## loading in the non OSNs from the occlusion
```{r}
#continue from here
occlusion_non_OSNs <- readRDS("/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/occlusion_non_OSNs.rds")
```


#duplicating them
```{r}
merged_non_OSNs_2 <- merged_non_OSNs
occlusion_non_OSNs_2 <- occlusion_non_OSNs


```


```{r}
merged_non_OSNs_2$status <- "wildtype"
occlusion_non_OSNs_2$status <- "occlusion"

```



```{r}
DefaultAssay(merged_non_OSNs_2) <- "RNA"
DefaultAssay(occlusion_non_OSNs_2) <- "RNA"

merged_non_OSNs_2[["SCT"]] <- NULL
occlusion_non_OSNs_2[["SCT"]] <- NULL

```






## merging the two non-OSN datasets together

```{r}
#merging the seurat objects

combined_non_OSNs <- merge(x = merged_non_OSNs_2, y = occlusion_non_OSNs_2)

```


```{r}
# Rerun SCTransform  
combined_non_OSNs <- SCTransform(combined_non_OSNs, verbose = FALSE)

#Run PCA and UMAP for visualization
combined_non_OSNs <- RunPCA(combined_non_OSNs, verbose = FALSE)
combined_non_OSNs <- RunUMAP(combined_non_OSNs, dims = 1:30)


```

```{r}
# Plot UMAP 
DimPlot(combined_non_OSNs, group.by = "orig.ident", label = TRUE) + NoLegend()


```





#Differential Expression Between wildtype and occlusion

```{r}
# Set SCT as default assay
DefaultAssay(combined_non_OSNs) <- "SCT"

# Prep for DE using SCT assay
combined_non_OSNs <- PrepSCTFindMarkers(combined_non_OSNs)

# Set the identity class
Idents(combined_non_OSNs) <- "status"

# Run DE between occlusion and wildtype
status.de.markers_non <- FindMarkers(combined_non_OSNs,
                                 ident.1 = "occlusion",
                                 ident.2 = "wildtype",
                                 logfc.threshold = 0.25,
                                 min.pct = 0.1,
                                 only.pos = FALSE,
                                 assay = "SCT")

# View results
head(status.de.markers_non)
```


other 0.25
```{r}
#Defining the significance thresholds
pval_threshold <- 0.05
logfc_threshold <- 0.25

# Subset dataframe for significant genes (adjusted p-value and log fold change)
significant_genes_df <- status.de.markers_non[
  status.de.markers_non$p_val_adj < pval_threshold & 
  abs(status.de.markers_non$avg_log2FC) > logfc_threshold, 
]

# View the first few significant genes
head(significant_genes_df)

```


## loading in data frame for significant differentially expressed genes from OSNs
```{r}
significant_OSN_genes_df <- readRDS("/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/significant_OSN_genes_df.rds")
```

```{r}
head(significant_OSN_genes_df)
```


```{r}
# Get gene names from both dataframes 
genes_in_significant_genes_df <- rownames(significant_genes_df)
genes_in_significant_OSN_genes_df <- rownames(significant_OSN_genes_df)

# Find genes unique to significant_OSN_genes_df
unique_genes_OSN <- setdiff(genes_in_significant_OSN_genes_df, genes_in_significant_genes_df)

# Subset significant_OSN_genes_df to keep only unique genes
unique_significant_OSN_df <- significant_OSN_genes_df[unique_genes_OSN, ]

# View the first few rows
head(unique_significant_OSN_df)

```


#creating the universe, gather genes of combined_non_OSNs and of harmony_nasal

loading in the harmony_nasal
```{r}

#continue from here
harmony_nasal <- readRDS("/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/harmony_nasal.rds")
```



#making vector of genes from harmony_nasal
```{r}
OSN_gene_list <- rownames(harmony_nasal)
head(OSN_gene_list)
```

##making vector of genes from the nonOSNs
```{r}
non_OSN_gene_list <- rownames(combined_non_OSNs)
head(non_OSN_gene_list)
```

##actually creating the universe vector, while removing duplicates
```{r}
universe_list <- unique(c(OSN_gene_list, non_OSN_gene_list))
```


## this entire workflow was to find differentially expressing genes that were only present in OSNs by looking at differentially expressing genes in non OSNs too, and then taking the unique differentially expressing genes in OSNs


```{r}

GO <- enrichGO(gene         = rownames(unique_significant_OSN_df),
                universe      = universe_list,
                OrgDb         = org.Mm.eg.db,
                keyType       = 'SYMBOL',
                ont           = "All",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.01,
                qvalueCutoff  = 0.05)
head(GO)    
```




```{r}
barplot(GO, showCategory = 10, title = "Gene Ontology Enrichment")
```
                          

```{r}
# Convert GO result to data frame
GO_df <- as.data.frame(GO)

# Take top 10 by p-value
top_GO <- GO_df %>%
  arrange(pvalue) %>%
  slice(1:10)

# Plot
ggplot(top_GO, aes(x = reorder(Description, -log10(pvalue)), y = -log10(pvalue), fill = -log10(pvalue))) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_gradient(low = "skyblue", high = "red") +
  labs(
    x = "GO Term",
    y = "-log10(p-value)",
    title = "GO Enrichment by p-value",
    fill = "-log10(p-value)"
  ) +
  theme_minimal()



```


```{r}
# Convert GO to dataframe 
GO_df <- as.data.frame(GO)

# Take top 10 by Fold Enrichment
top_GO <- GO_df %>%
  arrange(desc(FoldEnrichment)) %>%
  slice(1:10)

# Plot
ggplot(top_GO, aes(x = reorder(Description, FoldEnrichment), 
                   y = FoldEnrichment, 
                   fill = FoldEnrichment)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_gradient(low = "lightblue", high = "red") +
  labs(
    title = "Top GO Terms by Fold Enrichment",
    x = "GO Term",
    y = "Fold Enrichment",
    fill = "FoldEnrichment"
  ) +
  theme_minimal()
```


use significant non-osns to remove non-osns from volcano plot and the replot. Volcano plot should have significant and non-significant genes. From the nasal_occlusion_edits markdown

# dataframe of significant non-OSN genes
```{r}
head(significant_genes_df)
```






```{r}

status.de.markers <- readRDS("/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/status.de.markers.rds")
```


```{r}
head(status.de.markers)
```



```{r}#
status.de.markers %>% 
  rownames_to_column("Gene") %>%
  filter(!Gene %in% rownames(significant_genes_df))
```




```{r}
filtered_status.de.markers <-status.de.markers %>% 
  rownames_to_column("Gene") %>%
  filter(!Gene %in% rownames(significant_genes_df))
```

```{r}
head(filtered_status.de.markers)
```








```{r}
# Prepare data for volcano plot
volcano_data <- filtered_status.de.markers
volcano_data$gene <- filtered_status.de.markers$Gene

ggplot(volcano_data, aes(x = avg_log2FC, y = -log10(p_val_adj))) +
  geom_point(alpha = 0.5) +
  theme_minimal() +
  xlab("Average Log2 Fold Change") +
  ylab("-Log10 Adjusted P-value") +
  geom_vline(xintercept = c(-0.5, 0.5), col = "red", linetype = "dashed") +
  geom_hline(yintercept = -log10(0.05), col = "red", linetype = "dashed") +
  geom_text(data = subset(volcano_data, p_val_adj < 0.05 & abs(avg_log2FC) > 1),
            aes(label = gene), vjust = 1, hjust = 1, size = 3)

```

```{r}
# Add status column for color coding
volcano_data <- filtered_status.de.markers
volcano_data$gene <- volcano_data$Gene

volcano_data <- volcano_data %>%
  mutate(status = case_when(
    p_val_adj < 0.05 & avg_log2FC > 0.25 ~ "Upregulated",
    p_val_adj < 0.05 & avg_log2FC < -0.25 ~ "Downregulated",
    TRUE ~ "Not Significant"
  ))

# Volcano plot with color-coded points
ggplot(volcano_data, aes(x = avg_log2FC, y = -log10(p_val_adj), color = status)) +
  geom_point(alpha = 0.7) +
  scale_color_manual(values = c(
    "Upregulated" = "red",
    "Downregulated" = "steelblue",
    "Not Significant" = "grey"
  )) +
  theme_minimal() +
  xlab("Average Log2 Fold Change") +
  ylab("-Log10 Adjusted P-value") +
  geom_vline(xintercept = c(-0.5, 0.5), col = "black", linetype = "dashed") +
  geom_hline(yintercept = -log10(0.05), col = "black", linetype = "dashed") +
  geom_text(
    data = subset(volcano_data, p_val_adj < 0.05 & abs(avg_log2FC) > 1),
    aes(label = gene), vjust = 1, hjust = 1, size = 3, check_overlap = TRUE
  )

```


```{r}
# Volcano plot with color-coded points (no gene labels)
ggplot(volcano_data, aes(x = avg_log2FC, y = -log10(p_val_adj), color = status)) +
  geom_point(alpha = 0.7) +
  scale_color_manual(values = c(
    "Upregulated" = "red",
    "Downregulated" = "steelblue",
    "Not Significant" = "grey"
  )) +
  theme_minimal() +
  xlab("Average Log2 Fold Change") +
  ylab("-Log10 Adjusted P-value") +
  geom_vline(xintercept = c(-0.5, 0.5), col = "black", linetype = "dashed") +
  geom_hline(yintercept = -log10(0.05), col = "black", linetype = "dashed")

```






```{r}
AIS_categories <- read.csv("/scratch/prj/dn_hamidlab_msc_proj/k24082190/originals/New_AIS_cat_prop.csv")
```



```{r}
AIS_categories[12, 5] <- "Enzyme"
```




```{r}
AIS_categories <- AIS_categories %>%
  filter(!grepl("look above", AIS_specific, ignore.case = TRUE))

```

```{r}
AIS_categories <- AIS_categories %>%
  filter(!grepl("look below", AIS_specific, ignore.case = TRUE))
```


```{r}
new_AIS_genes_df <- as.data.frame(AIS_categories)
```


```{r}
keeps <- c("Mouse_name", "AIS_specific")
```


```{r}
new_AIS_genes_df <- new_AIS_genes_df[keeps]
```


```{r}
head(new_AIS_genes_df)
```



```{r}
new_AIS_genes_df <- new_AIS_genes_df[new_AIS_genes_df$AIS_specific %in% c("yes", "maybe", "Yes", "Maybe"), ]

```


```{r}
new_AIS_genes_df[17, 1] <- "Kcnb2"

```


```{r}
new_AIS_genes <- as.character(new_AIS_genes_df$Mouse_name)
```

```{r}
new_AIS_genes
```

```{r}
head(filtered_status.de.markers)
```


creates a new dataframe, filtered_status.de.markers_AIS, that only includes rows where Gene is in the new_AIS_genes vector.
```{r}
filtered_status.de.markers_AIS <- filtered_status.de.markers %>%
  filter(Gene %in% new_AIS_genes)
```

```{r}
head(filtered_status.de.markers_AIS)
```






#top 10 differentially expressed AIS genes
```{r}
# Top 10 genes by adjusted p-value
top_10_genes <- filtered_status.de.markers_AIS %>%
  arrange(p_val_adj) %>%
  slice_head(n = 10) %>%
  pull(Gene)

```


```{r}
# Create individual violin plots for each 
vplots <- lapply(top_10_genes, function(gene) {
  VlnPlot(harmony_nasal, features = gene, group.by = "status", pt.size = 0.9) +
    ggtitle(gene)+
    theme(legend.position = "none")
})


vplots[[1]]
vplots[[2]]
vplots[[3]]
vplots[[4]]
vplots[[5]]
vplots[[6]]
vplots[[7]]
vplots[[8]]
vplots[[9]]
vplots[[10]]

```



```{r}
wrap_plots(vplots[c(1, 3, 4, 5)], ncol = 2)


```


```{r}
head(filtered_status.de.markers)
```

## doing differential expression between the two

```{r}
filtered_status.de.markers <- filtered_status.de.markers %>%
  mutate(direction = ifelse(avg_log2FC > 0, "occlusion", "wildtype"))
```

```{r}
genes_up_in_occlusion <- filtered_status.de.markers %>%
  filter(direction == "occlusion", p_val_adj < 0.05) %>%
  pull(Gene)

genes_up_in_wildtype <- filtered_status.de.markers %>%
  filter(direction == "wildtype", p_val_adj < 0.05) %>%
  pull(Gene)

```

```{r}
# GO enrichment
ego_occlusion_filtered <- enrichGO(gene = genes_up_in_occlusion,
                          universe      = universe_list,         
                          OrgDb = org.Mm.eg.db,
                          keyType = "SYMBOL",
                          ont = "ALL",  
                          pAdjustMethod = "BH",
                          qvalueCutoff = 0.05,
                          readable = TRUE)

ego_wildtype_filtered <- enrichGO(gene = genes_up_in_wildtype,
                         universe      = universe_list,
                         OrgDb = org.Mm.eg.db,
                         keyType = "SYMBOL",
                         ont = "ALL",
                         pAdjustMethod = "BH",
                         qvalueCutoff = 0.05,
                         readable = TRUE)
```


```{r}
barplot(ego_occlusion_filtered, showCategory = 10, title = "GO Enrichment: Occlusion Upregulated")
barplot(ego_wildtype_filtered, showCategory = 10, title = "GO Enrichment: Wildtype Upregulated")

```


```{r}
df_occlusion <- as.data.frame(ego_occlusion_filtered)
df_wildtype <- as.data.frame(ego_wildtype_filtered)

```



```{r}
# Occlusion plot
ggplot(head(df_occlusion[order(-df_occlusion$FoldEnrichment), ], 10),
       aes(x = reorder(Description, FoldEnrichment), y = FoldEnrichment, fill = FoldEnrichment)) +
  geom_col() +
  coord_flip() +
  scale_fill_gradient(low = "lightblue", high = "red") +
  labs(title = "GO Enrichment: Occlusion Upregulated",
       x = "GO Term", y = "Fold Enrichment") +
  theme_minimal()

# Wildtype plot
ggplot(head(df_wildtype[order(-df_wildtype$FoldEnrichment), ], 10),
       aes(x = reorder(Description, FoldEnrichment), y = FoldEnrichment, fill = FoldEnrichment)) +
  geom_col() +
  coord_flip() +
  scale_fill_gradient(low = "lightblue", high = "red") +
  labs(title = "GO Enrichment: Wildtype Upregulated",
       x = "GO Term", y = "Fold Enrichment") +
  theme_minimal()
```


```{r}
# Occlusion plot
ggplot(
  head(df_occlusion[order(-df_occlusion$FoldEnrichment), ], 10),
  aes(x = reorder(Description, FoldEnrichment), y = FoldEnrichment, fill = FoldEnrichment)
) +
  geom_col() +
  coord_flip() +
  scale_fill_gradient(low = "lightblue", high = "red") +
  labs(title = "GO Enrichment: Occlusion Upregulated",
       x = "GO Term", y = "Fold Enrichment") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 14),
    axis.text.x = element_text(size = 14),
    axis.title = element_text(size = 13)
  )

# Wildtype plot clean label and enlarge text
df_wildtype_clean <- df_wildtype %>%
  mutate(Description = str_replace(
    Description,
    "protein insertion into ER membrane by stop-transfer membrane-anchor sequence",
    "protein insertion into ER membrane"
  ))

ggplot(
  head(df_wildtype_clean[order(-df_wildtype_clean$FoldEnrichment), ], 10),
  aes(x = reorder(Description, FoldEnrichment), y = FoldEnrichment, fill = FoldEnrichment)
) +
  geom_col() +
  coord_flip() +
  scale_fill_gradient(low = "lightblue", high = "red") +
  labs(title = "GO Enrichment: Wildtype Upregulated",
       x = "GO Term", y = "Fold Enrichment") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 14),
    axis.text.x = element_text(size = 14),
    axis.title = element_text(size = 13)
  )

```







































```{r}
# Ensure group labels exist (e.g., 'condition' column)
Idents(harmony_nasal_filtered) <- "condition"

# DE analysis
de_results <- FindMarkers(
  object = harmony_nasal_filtered,
  ident.1 = "occlusion",
  ident.2 = "wildtype",
  logfc.threshold = 0,
  min.pct = 0.1,
  test.use = "wilcox"
)

# Add gene names as a column
de_results$gene <- rownames(de_results)

```

















