---
title: "Nasal occlusion"
output: html_document
date: "2025-05-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/scratch/prj/dn_hamidlab_msc_proj/k24082190/originals")
```

```{r}
#Setting working directory
setwd("/scratch/prj/dn_hamidlab_msc_proj/k24082190/originals")
```

```{r}
#setting custom path
.libPaths("/scratch/prj/dn_hamidlab_msc_proj/Libraries")
```




```{r}
#load the libraries needed
## load dependencies
library(tidyverse)
library(Seurat)
library(SeuratObject)
library(sp)
library(scDblFinder)
library(SingleCellExperiment)
#library(SeuratWrappers)
#library(SeuratDisk)
library(patchwork)
library(clustree)
library(dplyr)
library(XML)
library(Matrix)
library(glmGamPoi)
library(gprofiler2)
library(harmony)
library(purrr)
library(limma)
library(umap)
library(pheatmap)
library(fgsea)
library(clusterProfiler)
library(org.Mm.eg.db)
library(rrvgo)
library(slingshot)
library(destiny)
library(tradeSeq)
library(EnhancedVolcano)
```

```{r}
# Set file paths
data_dir <- "/scratch/prj/dn_hamidlab_msc_proj/k24082190/originals/GSE157119_RAW"

matrix_file   <- file.path(data_dir, "GSE157119_aggregation_matrix.mtx.gz")
features_file <- file.path(data_dir, "GSE157119_aggregation_features.tsv.gz")
barcodes_file <- file.path(data_dir, "GSE157119_aggregation_barcodes.tsv.gz")

# Read the matrix using ReadMtx
nasal_counts <- ReadMtx(mtx = matrix_file,
                        features = features_file,
                        cells = barcodes_file)

# Create Seurat object
nasal_seurat <- CreateSeuratObject(counts = nasal_counts,
                                   project = "GSE157119")

# Label cells based on barcode suffix
nasal_seurat$condition <- ifelse(grepl("-1$", colnames(nasal_seurat)), "closed", "open")

# Preview
nasal_seurat
```


```{r}
# Convert to SingleCellExperiment
sce <- as.SingleCellExperiment(nasal_seurat)

# Run scDblFinder
sce <- scDblFinder(sce)

# Add doublet classifications back to Seurat object
nasal_seurat$scDblFinder.class <- sce$scDblFinder.class

```

#filtering out doublets
```{r}
nasal_seurat <- subset(nasal_seurat, subset = scDblFinder.class == "singlet")

```

#qc check for mitochondrial
```{r}
# Calculate the percentage of mitochondrial genes
nasal_seurat$percent_mito <- PercentageFeatureSet(nasal_seurat, pattern = "^mt-", assay = "RNA")

# View the first few entries of the 'percent_mito' column in the metadata
head(nasal_seurat@meta.data$percent_mito)

```

```{r}
# Visualize the mitochondrial gene percentage distribution across cells
VlnPlot(nasal_seurat, features = "percent_mito", pt.size = 0.1)

```


```{r}
# Calculate percent ribosomal genes for the nasal_seurat object
nasal_seurat$percent_ribo <- PercentageFeatureSet(nasal_seurat, pattern = "^Rp[sl]", assay = "RNA")

```


```{r}
# Plot ribosomal gene percentage distribution for the nasal_seurat object
VlnPlot(nasal_seurat, features = "percent_ribo", pt.size = 0.1) +
  ggtitle("Ribosomal Gene Percentage") +
  scale_y_continuous(limits = c(0, 65))  

```


```{r}
cc.genes <- cc.genes.updated.2019
```


```{r}

cc.genes <- lapply(cc.genes, function(genes) {
   gprofiler2::gorth(genes, 
        target_organism = "mmusculus"   
   )$ortholog_name
})
```

```{r}
head(cc.genes)
```

```{r}
nasal_seurat <- NormalizeData(nasal_seurat)
nasal_seurat <- CellCycleScoring(nasal_seurat,
                                  s.features = cc.genes$s.genes,
                                  g2m.features = cc.genes$g2m.genes,
                                  set.ident = FALSE)

```

```{r}
ggplot(nasal_seurat@meta.data, aes(x = Phase, fill = Phase)) +
  geom_bar() +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2") +
  ggtitle(unique(nasal_seurat$orig.ident)) +
  theme(legend.position = "none")
```

#regressing out cc score
```{r}
nasal_seurat <- SCTransform(nasal_seurat, 
                             method = "glmGamPoi", 
                             vars.to.regress = c("S.Score", "G2M.Score"), 
                             verbose = FALSE)
```


```{r}
# Check number of cells before QC
ncell_before <- ncol(nasal_seurat)
print(ncell_before)

```



```{r}
# 1. Calculate QC cutoffs using median and MAD
feature_cutoff <- median(nasal_seurat@meta.data$nFeature_SCT) - 3 * mad(nasal_seurat@meta.data$nFeature_SCT)
feature_cutoff <- ifelse(feature_cutoff < 100, 100, feature_cutoff)

mito_cutoff <- median(nasal_seurat@meta.data$percent_mito) + 3 * mad(nasal_seurat@meta.data$percent_mito)
mito_cutoff <- ifelse(mito_cutoff > 10, 10, mito_cutoff)

# Print the cutoffs
qc.df <- data.frame(feature_cutoff = feature_cutoff,
                    mito_cutoff = mito_cutoff)
print(qc.df)

# 2. Filter out poor-quality cells based on these thresholds
nasal_seurat <- subset(nasal_seurat,
                                subset = nFeature_SCT > feature_cutoff &
                                         percent_mito < mito_cutoff)




```

```{r}
print(nasal_seurat)
```



```{r}
# Number of cells after QC filtering
ncell_after <- ncol(nasal_seurat)
print(ncell_after)

```


```{r}
# Violin plot for percent mitochondrial genes
VlnPlot(nasal_seurat, features = "percent_mito") + 
  ggtitle("Percent Mitochondrial Genes")

# Violin plot for number of features (genes) detected
VlnPlot(nasal_seurat, features = "nFeature_RNA") + 
  ggtitle("Number of Features (RNA)")
```


```{r}
#normalising
nasal_seurat <- SCTransform(nasal_seurat,
                            method = "glmGamPoi",
                            verbose = FALSE)
```

```{r}
# Run PCA
nasal_seurat <- RunPCA(nasal_seurat, verbose = FALSE)

# Find neighbors using the first 50 PCs
nasal_seurat <- FindNeighbors(nasal_seurat, dims = 1:50)

# Find clusters with higher resolution (1.8)
nasal_seurat <- FindClusters(nasal_seurat, resolution = 1.8)

```



```{r}
# check cluster identities
table(Idents(nasal_seurat))

```

```{r}
# Violin plot for percent mitochondrial genes
p1 <- VlnPlot(nasal_seurat, features = "percent_mito") +
  ggtitle("Percent Mitochondrial Content")

# Violin plot for number of features (genes) detected
p2 <- VlnPlot(nasal_seurat, features = "nFeature_RNA") +
  ggtitle("Number of Features (nFeature_RNA)")

# Combine the two plots vertically
p1 + p2 + plot_layout(ncol = 1)
```


```{r}
# Identify clusters that do NOT meet the low-quality criteria
clusters_to_keep <- nasal_seurat@meta.data %>%
  group_by(seurat_clusters) %>%
  summarise(
    maxfeat = max(nFeature_RNA),
    maxmito = max(percent_mito)
  ) %>%
  filter(maxfeat > 1000 | maxmito < 2.5) %>%
  pull(seurat_clusters)

# Subset Seurat object to keep only good clusters
nasal_seurat <- subset(nasal_seurat, idents = clusters_to_keep)
```


```{r}
# Plot for percent mitochondrial content
p1 <- VlnPlot(nasal_seurat, features = "percent_mito") +
  ggtitle("Percent Mitochondrial Content")

# Plot for number of detected features (genes)
p2 <- VlnPlot(nasal_seurat, features = "nFeature_RNA") +
  ggtitle("Number of Features (nFeature_RNA)")

# Combine vertically
p1 + p2 + plot_layout(ncol = 1)
```

```{r}
# Re-run neighbors and clustering
nasal_seurat <- FindNeighbors(nasal_seurat, dims = 1:50)
nasal_seurat <- FindClusters(nasal_seurat, resolution = 1.8)

```

```{r}
# Identify clusters that pass QC thresholds
clusters_to_keep2 <- nasal_seurat@meta.data %>%
  group_by(seurat_clusters) %>%
  summarise(
    maxfeat = max(nFeature_RNA),
    maxcount = max(nCount_RNA)
  ) %>%
  filter(maxfeat > 2000 | maxcount > 4000) %>%
  pull(seurat_clusters)

# Subset to keep only those clusters
nasal_seurat <- subset(nasal_seurat, idents = clusters_to_keep2)
```



```{r}
# Violin plot for nCount_RNA
p1 <- VlnPlot(nasal_seurat, features = "nCount_RNA") +
  ggtitle("nCount_RNA Distribution")

# Violin plot for nFeature_RNA
p2 <- VlnPlot(nasal_seurat, features = "nFeature_RNA") +
  ggtitle("nFeature_RNA Distribution")

# Combine plots vertically
p1 + p2 + plot_layout(ncol = 1)
```

#Final SCT nasal_seurat
```{r}
nasal_seurat <- SCTransform(nasal_seurat,
                            method = "glmGamPoi",
                            verbose = FALSE)
```

#saving rds file
```{r}
#saveRDS(nasal_seurat, file = "/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/FinalSCT_nasal_seurat.rds")
```


```{r}
### from here load in this rds and continue
nasal_seurat <- readRDS("/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/FinalSCT_nasal_seurat.rds")
```


```{r}
# Run dimensionality reduction on nasal_seurat
nasal_seurat <- RunPCA(nasal_seurat)
nasal_seurat <- RunTSNE(nasal_seurat, dims = 1:50)
nasal_seurat <- RunUMAP(nasal_seurat, dims = 1:50)

```

```{r}
DimPlot(nasal_seurat, reduction = "umap")

```

```{r}
DimPlot(nasal_seurat, reduction = "tsne")

```


```{r}
# Set range of clustering resolutions
res.range <- seq(0.4, 3, 0.2)

# Set default assay and find neighbors
DefaultAssay(nasal_seurat) <- "SCT"
nasal_seurat <- FindNeighbors(nasal_seurat, dims = 1:50)

# Loop through resolutions and cluster
for (res in res.range) {
  nasal_seurat <- FindClusters(nasal_seurat, resolution = res)
}

```

```{r}
clustree::clustree(nasal_seurat, prefix = "SCT_snn_res.")
```


```{r}


wss.out <- do.call(bind_rows, lapply(res.range, function(k) {
  
  # Get cluster IDs for resolution k
  cluster_col <- paste0("SCT_snn_res.", k)
  
  # Calculate centroids for each cluster in PCA space
  centroids <- aggregate(Embeddings(nasal_seurat, reduction = "pca"),
                         by = list(cluster = nasal_seurat[[cluster_col]][,1]),
                         FUN = mean)
  rownames(centroids) <- centroids$cluster
  centroids <- centroids[, -1]
  
  # Calculate squared differences between each cell and its cluster centroid
  diffsq <- (Embeddings(nasal_seurat, reduction = "pca") - 
               centroids[nasal_seurat@meta.data[[cluster_col]], ])^2
  
  # Calculate variance (sd squared) for each PC
  sdsq <- colSds(Embeddings(nasal_seurat, reduction = "pca"))^2
  
  # Weight differences by variance
  wdiff <- diffsq / sdsq
  
  # Calculate weighted sum of squares
  wss <- sum(sqrt(rowSums(wdiff)))
  
  data.frame(clust = k, wss = wss)
}))

# Add sample ID
wss.out$id <- nasal_seurat$orig.ident[1]

wss.out

```

```{r}
wss.out %>%
  mutate(scaled_wss = scale(wss)) %>%
  ggplot(aes(x = clust, y = scaled_wss)) +
  geom_line(color = "blue") +
  geom_point(color = "blue") +
  labs(title = paste("WSS vs Clustering Resolution for", wss.out$id[1]),
       x = "Resolution",
       y = "Scaled Weighted Sum of Squares") +
  theme_minimal()
```


```{r}


# Define to test
resolutions <- c(0.8, 1.2, 1.6, 2.0, 2.4, 2.6)

# Run clustering for each resolution and store cluster identities in metadata
for (res in resolutions) {
  nasal_seurat <- FindClusters(nasal_seurat, resolution = res, verbose = FALSE)
  # creates metadata columns like "SCT_snn_res.1.2", etc.
}

# Plot UMAPs for each resolution side by side
plots <- lapply(resolutions, function(res) {
  cluster_col <- paste0("SCT_snn_res.", res)
  Idents(nasal_seurat) <- nasal_seurat[[cluster_col]][,1]  # extract vector properly
  DimPlot(nasal_seurat, reduction = "umap", label = TRUE) + 
    ggtitle(paste("Resolution", res))
})

# Combine plots
wrap_plots(plots, ncol = 2)

```

```{r}
plots <- lapply(resolutions, function(res) {
  cluster_col <- paste0("SCT_snn_res.", res)
  Idents(nasal_seurat) <- nasal_seurat@meta.data[[cluster_col]]
  DimPlot(nasal_seurat, reduction = "umap", label = TRUE) + 
    ggtitle(paste("Resolution", res))
})

```



```{r}
print(plots[[1]]) 
print(plots[[2]])  
print(plots[[3]]) 
print(plots[[4]]) 
print(plots[[5]]) 

```


```{r}
## clustering for real

res <- 1.2
nasal_seurat <- FindNeighbors(nasal_seurat, verbose = F)
nasal_seurat <- FindClusters(nasal_seurat, resolution = res, verbose = F)

```


```{r}
Idents(nasal_seurat) <- "SCT_snn_res.1.2"
FeaturePlot(nasal_seurat, features = c("Syn1", "Tubb3", "Gap43"))

```



```{r}
table(nasal_seurat$seurat_clusters)
```

```{r}
DotPlot(nasal_seurat, features = c("Syn1" ), split.by = NULL)
```


#saving rds file
```{r}
#saveRDS(nasal_seurat, file = "/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/FinalSCT_nasal_seurat.rds")
```


```{r}
### from here load in this rds and continue::::::::
nasal_seurat <- readRDS("/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/FinalSCT_nasal_seurat.rds")
```



```{r}
#importing mouse genes
mouse_genes = read.csv("/scratch/prj/dn_hamidlab_msc_proj/k24082190/originals/Dishalas_Filtered_OSN_Genes.csv")
```


```{r}
#importing human genes
human_genes = read.csv("/scratch/prj/dn_hamidlab_msc_proj/k24082190/originals/Filtered_human_genes.csv")
```


```{r}
# Filter the data for mOSN (mature OSN) cluster
mouse_mOSN_info = mouse_genes[mouse_genes$cluster == "mOSN", ]

# Filter the data for iOSN (immature OSN) cluster
mouse_iOSN_info = mouse_genes[mouse_genes$cluster == "iOSN", ]
```


```{r}
human_gene_lists = split(human_genes$gene, human_genes$cluster)

# Separate into iOSN and mOSN genes
human_mOSN_genes = human_genes[human_genes$cluster =="Mature Neurons", ]
human_iOSN_genes = human_genes[human_genes$cluster =="Immature Neurons", ]


# Convert human mOSN genes to mouse orthologs
converted_human_mOSN_genes = gprofiler2::gorth(
  human_mOSN_genes$gene, 
  source_organism = "hsapiens", 
  target_organism = "mmusculus"
)

# Convert human iOSN genes to mouse orthologs
converted_human_iOSN_genes = gprofiler2::gorth(
  human_iOSN_genes$gene, 
  source_organism = "hsapiens", 
  target_organism = "mmusculus"
) 
```



```{r}
# Extract only necessary columns
converted_human_mOSN_df <- converted_human_mOSN_genes[, c("input", "ortholog_name")]
converted_human_iOSN_df <- converted_human_iOSN_genes[, c("input", "ortholog_name")]

# Renaming the column nmes
colnames(converted_human_mOSN_df) <- c("gene", "mouse_gene")
colnames(converted_human_iOSN_df) <- c("gene", "mouse_gene")

# Merge back with original data to preserve other columns like p-value, avg_logFC and so on
human_mOSN_info <- merge(human_mOSN_genes, converted_human_mOSN_df, 
                         by = "gene", all.x = TRUE)

human_iOSN_info <- merge(human_iOSN_genes, converted_human_iOSN_df, 
                         by = "gene", all.x = TRUE)

# Add source column
human_mOSN_info$source <- "converted_human_mOSN"
human_iOSN_info$source <- "converted_human_iOSN"

# View results
head(human_mOSN_info)
head(human_iOSN_info)





```



```{r}

# Standardisinge column names by ensuring all data frames have the same structure
human_iOSN_info$species <- "Human"
human_mOSN_info$species <- "Human"
mouse_mOSN_info$species <- "Mouse"
mouse_iOSN_info$species <- "Mouse"

# Add missing columns (source in mouse dataset)
mouse_mOSN_info$source <- "converted_mouse_mOSN"
mouse_iOSN_info$source <- "converted_mouse_iOSN"

# Making sure all have the same column order
all_columns <- c("gene", "p_val", "avg_logFC", "pct.1", "pct.2", "p_val_adj", "cluster", "mouse_gene", "source", "species")

# Add missing columns to mouse data frames
mouse_mOSN_info$mouse_gene <- mouse_mOSN_info$gene
mouse_iOSN_info$mouse_gene <- mouse_iOSN_info$gene

# Bind all data frames
combined_df <- bind_rows(
  dplyr::select(mouse_mOSN_info, all_of(all_columns)),  
  dplyr::select(mouse_iOSN_info, all_of(all_columns)),
  dplyr::select(human_iOSN_info, all_of(all_columns)),  
  dplyr::select(human_mOSN_info, all_of(all_columns))
)

# View the combined data frame
head(combined_df)



```


```{r}
combined_df %>%
  distinct(mouse_gene, .keep_all = T) %>%   ##this line gets the distinct genes, gets rid of human duplicate
  pull(cluster) %>% unique()
```

```{r}
# Recode human clusters to match mouse clusters
combined_df <- combined_df %>%
  mutate(cluster = case_when(
    cluster == "Immature Neurons" ~ "iOSN",  # Recode "Immature Neurons" to "iOSN"
    cluster == "Mature Neurons" ~ "mOSN",    # Recode "Mature Neurons" to "mOSN"
    TRUE ~ cluster                          # Keep other clusters unchanged
  ))
```


```{r}
combined_df %>%
  group_by(cluster) %>%
  slice_max(n=100, order_by = avg_logFC) %>%  ##getting the top 100 for mOSN and iOSN each according to avg_logFC
  group_by(species, cluster) %>%
  tally()
```


#extracting iOSN and mOSN genes separately
```{r}
top_mOSN_genes <- combined_df %>%
  filter(cluster == "mOSN") %>%
  slice_max(n = 100, order_by = avg_logFC) %>%
  pull(mouse_gene)

top_iOSN_genes <- combined_df %>%
  filter(cluster == "iOSN") %>%
  slice_max(n = 100, order_by = avg_logFC) %>%
  pull(mouse_gene)



```




```{r}
# Store in a list for AddModuleScore
top_osn_gene_sets <- list("iOSN" = top_iOSN_genes, "mOSN" = top_mOSN_genes)
```




#doing add module score
```{r}
nasal_seurat <- AddModuleScore(object = nasal_seurat,
    features = top_osn_gene_sets,  # List of gene sets
    name = "TopOSN_Score",  
    ctrl = 100, 
    assay = "RNA"  )
```

#score 1 corresponds to immature, score 2 corresponds to mature



```{r}
FeaturePlot(
  nasal_seurat,
  features = c("TopOSN_Score1","TopOSN_Score2") 
  
)
```

in no occlusion dataset the immature (score1) and mature (score2) cells separate cleanly, but in nasal occlusion dataset they overlap more, likely reflects a biological effect of the occlusion. It might be causing:

Delayed or disrupted OSN maturation

A more continuous or mixed state between immature and mature OSNs

A shift in gene expression dynamics, blurring the typical marker distinctions






```{r}
nasal_seurat@meta.data
```



```{r}


# Set cluster identities to "SCT_snn_res.1.2"
Idents(nasal_seurat) <- nasal_seurat$SCT_snn_res.1.2

# Feature plots for TopOSN_Score1 and TopOSN_Score2 
FeaturePlot(nasal_seurat, 
            features = "TopOSN_Score1", 
            label = TRUE, 
            min.cutoff = 0)

FeaturePlot(nasal_seurat, 
            features = "TopOSN_Score2", 
            label = TRUE, 
            min.cutoff = 0)

# Summarise TopOSN scores per cluster
nasal_seurat@meta.data %>%
  group_by(SCT_snn_res.1.2) %>%
  summarise(
    TopOSN_Score1_avg = mean(TopOSN_Score1, na.rm = TRUE),
    TopOSN_Score1_exp_perc = sum(TopOSN_Score1 > 0.1, na.rm = TRUE) / n(),
    TopOSN_Score2_avg = mean(TopOSN_Score2, na.rm = TRUE),
    TopOSN_Score2_exp_perc = sum(TopOSN_Score2 > 0.1, na.rm = TRUE) / n()
  )





```


```{r}
selected_clusters <- c(1, 4, 18, 21, 10, 11, 6, 25)

nasal_seurat_selected <- subset(nasal_seurat, idents = selected_clusters)

```



no need to run harmony or cca or rPCA because data is from one single experiment.

```{r}
ElbowPlot(nasal_seurat, ndims = 50)
```


###### nasal_seurat has all the clusters and nasal_seurat_selected has only the selected clusters. so you have two seurat objects at the moment.




```{r}
nasal_seurat_selected <- RunPCA(nasal_seurat_selected, verbose = FALSE)

# Find neighbors based on PCA
nasal_seurat_selected <- FindNeighbors(nasal_seurat_selected, dims = 1:50)

# Find clusters 
nasal_seurat_selected <- FindClusters(nasal_seurat_selected, resolution = 1.2)

# Visualise clusters using UMAP
nasal_seurat_selected <- RunUMAP(nasal_seurat_selected, dims = 1:50)
DimPlot(nasal_seurat_selected, reduction = "umap", label = TRUE, group.by = "orig.ident")

```

#OSN scores
```{r}
FeaturePlot(nasal_seurat_selected, features = "TopOSN_Score1", min.cutoff = 0, label = TRUE)  # iOSN
FeaturePlot(nasal_seurat_selected, features = "TopOSN_Score2", min.cutoff = 0, label = TRUE)  # mOSN

```


# osn_type to give a harder cutoff, more distinct classification
```{r}
nasal_seurat_selected$OSN_type <- ifelse(nasal_seurat_selected$TopOSN_Score1 > 
                                          nasal_seurat_selected$TopOSN_Score2, 
                                          "iOSN", "mOSN")

DimPlot(nasal_seurat_selected, reduction = "umap", group.by = "OSN_type", label = TRUE)

```


#saving rds file
```{r}
#saveRDS(nasal_seurat_selected, file = "/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/nasal_seurat_selected.rds")
```


```{r}
### from here load in this rds and continue
nasal_seurat_selected <- readRDS("/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/nasal_seurat_selected.rds")
```



# just checking for this gene
```{r}
FeaturePlot(
  nasal_seurat_selected,
  features = c("Tubb3"),
  label = TRUE
  
)
```

```{r}
FeaturePlot(
  nasal_seurat_selected,
  features = c("TopOSN_Score1", "TopOSN_Score2"),
  label = TRUE
  
)

```

```{r}
DimPlot(
  nasal_seurat_selected,
  group.by = "seurat_clusters",
  label = TRUE
)

```


```{r}
nasal_seurat_selected <- subset(nasal_seurat_selected, idents = c("0", "1", "2", "3", "4", "5", "6", "7", "9", "10"))

```

```{r}
saveRDS(nasal_seurat_selected, file = "/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/nasal_seurat_selected.rds")
```




#going to use the csv file that has the list of genes sorted into categories.

```{r}
AIS_categories <- read.csv("/scratch/prj/dn_hamidlab_msc_proj/k24082190/originals/New_AIS_cat_prop.csv")
```


```{r}
tail(AIS_categories)
```



```{r}
AIS_categories <- AIS_categories %>%
  filter(!grepl("look above", AIS_specific, ignore.case = TRUE))

```

```{r}
AIS_categories <- AIS_categories %>%
  filter(!grepl("look below", AIS_specific, ignore.case = TRUE))
```


```{r}
tail(AIS_categories)
```


```{r}
AIS_categories_df <- as.data.frame(AIS_categories)
```



```{r}
AIS_categories_df <- AIS_categories_df[, -1]

```


```{r}
AIS_categories_df[17, 1] <- "Kcnb2"

```

























































# getting the new list of AIS genes
```{r}
new_AIS_genes_csv <- read.csv("/scratch/prj/dn_hamidlab_msc_proj/k24082190/originals/New_AIS_cat_prop.csv")
```

```{r}
tail(new_AIS_genes_csv)
```

```{r}
new_AIS_genes_csv <- new_AIS_genes_csv %>%
  filter(!grepl("look above", AIS_specific, ignore.case = TRUE))

```

```{r}
new_AIS_genes_csv <- new_AIS_genes_csv %>%
  filter(!grepl("look below", AIS_specific, ignore.case = TRUE))
```


```{r}
new_AIS_genes_csv_df <- as.data.frame(new_AIS_genes_csv)
```


```{r}
keeps <- c("Mouse_name", "AIS_specific", "Function", "Category")
```


```{r}
new_AIS_genes_csv_df <- new_AIS_genes_csv_df[keeps]
```




```{r}
tail(new_AIS_genes_csv_df)
```

```{r}
new_AIS_genes_csv_df <- new_AIS_genes_csv_df[new_AIS_genes_csv_df$AIS_specific %in% c("yes", "maybe", "Yes", "Maybe"), ]

```

```{r}
new_AIS_genes <- as.character(new_AIS_genes_csv_df$Mouse_name)
```

```{r}
new_AIS_genes
```


















































`

























negative log10 of the p-values to better highlight statistical significance.
higher significance (lower p-values) as red and lower significance as blue in the gradient.


## doing differential expression again but with obj_harmony and nasal_selected_clusters


```{r}
# loading in obj_harmony

obj_harmony <- readRDS("/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/obj_harmony.rds")

```


```{r}
### from here load in this rds and continue
nasal_seurat_selected <- readRDS("/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/nasal_seurat_selected.rds")
```

```{r}
#making a copy of obj_harmony for merging

obj_harmony_2 <- obj_harmony
nasal_seurat_selected_2 <- nasal_seurat_selected
```



```{r}
obj_harmony_2$status <- "wildtype"
nasal_seurat_selected_2$status <- "occlusion"

```







```{r}
DefaultAssay(nasal_seurat_selected_2) <- "RNA"
DefaultAssay(obj_harmony_2) <- "RNA"

nasal_seurat_selected_2[["SCT"]] <- NULL
obj_harmony_2[["SCT"]] <- NULL

```


```{r}
shared_genes <- intersect(rownames(nasal_seurat_selected_2), rownames(obj_harmony_2))
length(shared_genes)  # should be > 0

```



```{r}
nasal_seurat_selected_2 <- nasal_seurat_selected_2[shared_genes, ]
obj_harmony_2 <- obj_harmony_2[shared_genes, ]

```







```{r}
#merging the seurat objects

harmony_nasal <- merge(x = obj_harmony_2, y = list(nasal_seurat_selected_2))
```


```{r}
head(harmony_nasal)
tail(harmony_nasal)
```


```{r}
# Rerun SCTransform  
harmony_nasal <- SCTransform(harmony_nasal, verbose = FALSE)

#Run PCA and UMAP for visualization
harmony_nasal <- RunPCA(harmony_nasal, verbose = FALSE)
harmony_nasal <- RunUMAP(harmony_nasal, dims = 1:30)


```


```{r}
#saveRDS(harmony_nasal, file = "/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/harmony_nasal.rds")

#continue from here
harmony_nasal <- readRDS("/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/harmony_nasal.rds")
```


```{r}
# Plot UMAP to check if immature and mature neurons separate well
DimPlot(harmony_nasal, group.by = "status", label = TRUE) + NoLegend()
```

```{r}
# Plot UMAP to check if immature and mature neurons separate well
DimPlot(harmony_nasal, group.by = "status", label = TRUE) + NoLegend()
```




```{r}
# Plot PCA by status 
DimPlot(harmony_nasal, reduction = "pca", group.by = "status", label = TRUE) + NoLegend()

```

```{r}
FeaturePlot(harmony_nasal, features = c("TopOSN_Score1", "TopOSN_Score2"))

```





#Differential Expression Between wildtype and occlusion

```{r}
# Set SCT as default assay
DefaultAssay(harmony_nasal) <- "SCT"

# Prep for DE using SCT assay
harmony_nasal <- PrepSCTFindMarkers(harmony_nasal)

# Set the identity class
Idents(harmony_nasal) <- "status"

# Run DE between occlusion and wildtype
status.de.markers <- FindMarkers(harmony_nasal,
                                 ident.1 = "occlusion",
                                 ident.2 = "wildtype",
                                 logfc.threshold = 0.25,
                                 min.pct = 0.1,
                                 only.pos = FALSE,
                                 assay = "SCT")

# View results
head(status.de.markers)
```

```{r}
saveRDS(status.de.markers, file = "/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/status.de.markers.rds")
```



data frame containing genes that are differentially expressed between the two status groups (occlusion vs wildtype)

p_val: raw p-value (unadjusted)

avg_log2FC: average log<sub>2</sub>-fold change between the two groups
Positive = up in occlusion
Negative = up in wildtype


pct.1: % of cells expressing the gene in occlusion

pct.2: % of cells expressing the gene in wildtype

p_val_adj: p-value adjusted for multiple testing (Benjamini-Hochberg)


```{r}
#Defining the significance thresholds
pval_threshold <- 0.05
logfc_threshold <- 0.5

# Subset dataframe for significant genes (adjusted p-value and log fold change)
significant_OSN_genes_df <- status.de.markers[
  status.de.markers$p_val_adj < pval_threshold & 
  abs(status.de.markers$avg_log2FC) > logfc_threshold, 
]

# View the first few significant genes
head(significant_OSN_genes_df)

```

```{r}
saveRDS(significant_OSN_genes_df, file = "/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/significant_OSN_genes_df.rds")
```








## violin plots
```{r}
# Top 10 genes by adjusted p-value
top_10_genes <- rownames(head(status.de.markers[order(status.de.markers$p_val_adj), ], 10))

# Violin plots for these genes
VlnPlot(harmony_nasal, features = top_10_genes, group.by = "status", pt.size = 0.9)

```


```{r}
# Create individual violin plots for each 
vplots <- lapply(top_10_genes, function(gene) {
  VlnPlot(harmony_nasal, features = gene, group.by = "status", pt.size = 0.9) +
    ggtitle(gene)  
})


vplots[[1]]
vplots[[2]]
vplots[[3]]
vplots[[4]]
vplots[[5]]
vplots[[6]]
vplots[[7]]
vplots[[8]]
vplots[[9]]
vplots[[10]]

```






```{r}

FeaturePlot(harmony_nasal, features = top_10_genes, reduction = "umap")

```



```{r}
# Prepare data for volcano plot
volcano_data <- status.de.markers
volcano_data$gene <- rownames(volcano_data)

ggplot(volcano_data, aes(x = avg_log2FC, y = -log10(p_val_adj))) +
  geom_point(alpha = 0.5) +
  theme_minimal() +
  xlab("Average Log2 Fold Change") +
  ylab("-Log10 Adjusted P-value") +
  geom_vline(xintercept = c(-0.5, 0.5), col = "red", linetype = "dashed") +
  geom_hline(yintercept = -log10(0.05), col = "red", linetype = "dashed") +
  geom_text(data = subset(volcano_data, p_val_adj < 0.05 & abs(avg_log2FC) > 1),
            aes(label = gene), vjust = 1, hjust = 1, size = 3)

```







#Perform DE analysis within the same cell type across conditions


OSN_type had na values for wildtype, not annotated, so here im annotating them (iosn and mosn_)
```{r}
na_idx <- which(is.na(harmony_nasal$OSN_type) & harmony_nasal$status == "wildtype")

harmony_nasal$OSN_type[na_idx] <- ifelse(
  harmony_nasal$TopOSN_Score1[na_idx] > harmony_nasal$TopOSN_Score2[na_idx],
  "iOSN",
  ifelse(
    harmony_nasal$TopOSN_Score2[na_idx] > harmony_nasal$TopOSN_Score1[na_idx],
    "mOSN",
    "Undefined"
  )
)

```


```{r}
harmony_nasal$celltype_status <- paste(harmony_nasal$OSN_type, harmony_nasal$status, sep = "_")
Idents(harmony_nasal) <- "celltype_status"

```


```{r}
table(harmony_nasal$celltype_status)

```

```{r}
# Make sure SCT is the default assay
DefaultAssay(harmony_nasal) <- "SCT"

# Prepare SCT for DE testing
harmony_nasal <- PrepSCTFindMarkers(harmony_nasal)
```


#Run DE for iOSN (wildtype vs occlusion):
```{r}
iosn.de <- FindMarkers(harmony_nasal,
                       ident.1 = "iOSN_occlusion",
                       ident.2 = "iOSN_wildtype",
                       logfc.threshold = 0.25,
                       min.pct = 0.1,
                       only.pos = FALSE)
head(iosn.de)

```

Xist-Extremely upregulated in occlusion: avg_log2FC = 11.7. Expressed in 86% of occlusion cells, 0% of wildtype. 

Rpl10-Downregulated in occlusion: avg_log2FC = -2.2. 60.8% vs 94.5% expression (occlusion vs wildtype)
Suggests lower ribosomal activity in iOSNs under occlusion

AC160336.1- upregulated: avg_log2FC = 7.99, expressed in 61% vs 0.5%
Could be a novel or less-characterized gene responsive to the condition






# Run DE for mOSN (wildtype vs occlusion):
```{r}
mosn.de <- FindMarkers(harmony_nasal,
                       ident.1 = "mOSN_occlusion",
                       ident.2 = "mOSN_wildtype",
                       logfc.threshold = 0.25,
                       min.pct = 0.1,
                       only.pos = FALSE)
head(mosn.de)

```

```{r}
AIS_categories <- read.csv("/scratch/prj/dn_hamidlab_msc_proj/k24082190/originals/New_AIS_cat_prop.csv")
 AIS_categories <- AIS_categories %>%
   filter(!grepl("look above", AIS_specific, ignore.case = TRUE))
 
 AIS_categories <- AIS_categories %>%
   filter(!grepl("look below", AIS_specific, ignore.case = TRUE))
 AIS_categories_df <- as.data.frame(AIS_categories)
 AIS_categories_df <- AIS_categories_df[, -1]
 
AIS_categories_df[17, 1] <- "Kcnb2"
```

```{r}
# Extract gene symbols as a vector
AIS_genes_violinlist <- AIS_categories_df$Mouse_name
```



```{r}
head(AIS_categories_df)
```


## for global


```{r}
# Keep only DE genes that are in the AIS list
AIS_sig_genes_df <- significant_OSN_genes_df[
  rownames(significant_OSN_genes_df) %in% AIS_genes_violinlist, 
]

```


```{r}
# Get top 10 by absolute log2FC
top10_AIS_genes <- AIS_sig_genes_df %>%
  dplyr::arrange(desc(abs(avg_log2FC))) %>%
  head(10)

# Gene names to plot
top10_gene_names <- rownames(top10_AIS_genes)

```

```{r}
top10_gene_names[!top10_gene_names %in% rownames(harmony_nasal[["SCT"]]@data)]

```






```{r}
VlnPlot(
  harmony_nasal,
  features = top10_gene_names,     # Use 'status' to group (e.g., wildtype vs occlusion)
  split.by = "status",     # Split violins by status within each gene
  pt.size = 0.1,
  ncol = 2,
  assay = "SCT"            # Or use "RNA" depending on what you normalized/used
)

```

```{r}
# Relevel the status factor to switch order
harmony_nasal$status <- factor(harmony_nasal$status, levels = c("occlusion", "wildtype"))

```






```{r}


# Set the identity to status so x-axis groups by status
Idents(harmony_nasal) <- "status"

# Now loop and plot with proper grouping
for (gene in top10_gene_names) {
  print(
    VlnPlot(
      object = harmony_nasal,
      features = gene,
      pt.size = 0.1,
      assay = "SCT"
    ) + ggtitle(gene)
  )
}


```



## results from most of the code below will not be used












## subset the three data frames for the ais genes.



## for iosn
```{r}
iosn.de$Gene <- rownames(iosn.de)

# Filter to only AIS genes
ais_genes_list <- AIS_categories_df$Mouse_name
iosn.de.ais <- iosn.de[iosn.de$Gene %in% ais_genes_list, ]

#  Merge DE results with AIS metadata
annotated_iosn_ais_de <- merge(
  iosn.de.ais,
  AIS_categories_df,
  by.x = "Gene",
  by.y = "Mouse_name"
)

head(annotated_iosn_ais_de)
```


################## new one 

```{r}


# Subset upregulated in occlusion (logFC > 0)
up_occlusion_genes <- annotated_iosn_ais_de %>%
  filter(avg_log2FC > 0.25) %>%
  pull(Gene)

# Subset upregulated in wildtype (logFC < 0)
up_wildtype_genes <- annotated_iosn_ais_de %>%
  filter(avg_log2FC < -0.25) %>%
  pull(Gene)

# Background genes = all genes tested
background_genes <- rownames(iosn.de)

# Run enrichment
ego_occlusion <- enrichGO(
  gene          = up_occlusion_genes,
  universe      = background_genes,
  OrgDb         = org.Mm.eg.db,
  keyType       = "SYMBOL",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.1,
  qvalueCutoff  = 0.1
)

ego_wildtype <- enrichGO(
  gene          = up_wildtype_genes,
  universe      = background_genes,
  OrgDb         = org.Mm.eg.db,
  keyType       = "SYMBOL",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.1,
  qvalueCutoff  = 0.1
)


```


```{r}
# Convert results to data frames
ego_occlusion_df <- as.data.frame(ego_occlusion)
ego_wildtype_df <- as.data.frame(ego_wildtype)

# View
head(ego_occlusion_df)
head(ego_wildtype_df)

```

```{r}
ego_occlusion_df$Group <- "Occlusion"
ego_wildtype_df$Group <- "Wildtype"

combined_df <- rbind(ego_occlusion_df, ego_wildtype_df)

# Select top 10 from each group
top_combined <- combined_df %>%
  group_by(Group) %>%
  slice_max(order_by = Count, n = 10, with_ties = FALSE) %>%
  ungroup()

ggplot(top_combined, aes(x = reorder(Description, Count), y = Count, fill = Group)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  labs(title = "Top GO Terms by Condition", x = "GO Term", y = "Gene Count") +
  theme_minimal()

```

```{r}
# Top 10 enriched GO terms 
top_terms <- ego_occlusion_df[1:10, ]

ggplot(top_terms, aes(x = reorder(Description, p.adjust), y = p.adjust, fill = p.adjust)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_gradient(low = "blue", high = "red") +
  labs(
    title = "Top 10 Enriched GO Terms for iOSNs (Occlusion)",
    x = "GO Term",
    y = "p.adjust",
    fill = "p.adjust"
  ) +
  theme_minimal()

```


```{r}
# Top 10 enriched GO terms 
top_terms <- ego_occlusion_df[1:10, ]

ggplot(top_terms, aes(x = reorder(Description, pvalue), y = pvalue, fill = pvalue)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_gradient(low = "blue", high = "red") +
  labs(
    title = "Top 10 Enriched GO Terms for iOSNs (Occlusion)",
    x = "GO Term",
    y = "pvalue",
    fill = "pvalue"
  ) +
  theme_minimal()

```


```{r}
# Top 10 enriched GO terms 
top_terms <- ego_occlusion_df[1:10, ]

ggplot(top_terms, aes(x = reorder(Description, FoldEnrichment), y = pvalue, fill = FoldEnrichment)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_gradient(low = "blue", high = "red") +
  labs(
    title = "Top 10 Enriched GO Terms for iOSNs (Occlusion)",
    x = "GO Term",
    y = "FoldEnrichment",
    fill = "FoldEnrichment"
  ) +
  theme_minimal()

```













```{r}
# Top 10 enriched GO terms (adjust as needed)
top_terms <- ego_wildtype_df[1:10, ]

ggplot(top_terms, aes(x = reorder(Description, p.adjust), y = p.adjust, fill = p.adjust)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_gradient(low = "blue", high = "red") +
  labs(
    title = "Top 10 Enriched GO Terms for iOSNs (Wildtype)",
    x = "GO Term",
    y = "p.adjust",
    fill = "p.adjust"
  ) +
  theme_minimal()

```

```{r}
# Top 10 enriched GO terms (adjust as needed)
top_terms <- ego_wildtype_df[1:10, ]

ggplot(top_terms, aes(x = reorder(Description, pvalue), y = pvalue, fill = pvalue)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_gradient(low = "blue", high = "red") +
  labs(
    title = "Top 10 Enriched GO Terms for iOSNs (Wildtype)",
    x = "GO Term",
    y = "pvalue",
    fill = "pvalue"
  ) +
  theme_minimal()

```
#### you stopped here
```{r}
# Top 10 enriched GO terms 
top_terms <- ego_wildtype_df[1:10, ]

ggplot(top_terms, aes(x = reorder(Description, FoldEnrichment), y = FoldEnrichment, fill = FoldEnrichment)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_gradient(low = "blue", high = "red") +
  labs(
    title = "Top 10 Enriched GO Terms for iOSNs (Wildtype)",
    x = "GO Term",
    y = "FoldEnrichment",
    fill = "FoldEnrichment"
  ) +
  theme_minimal()

```









```{r}


# Extract the gene list 
gene_list <- annotated_iosn_ais_de$Gene

# Define background/universe genes (all genes tested in iosn.de)
background_genes_i <- rownames(iosn.de)  # 

# Run GO enrichment
ego_ais_i <- enrichGO(
  gene          = gene_list,
  universe      = background_genes_i,
  OrgDb         = org.Mm.eg.db,
  keyType       = "SYMBOL",
  ont           = "ALL",          # or "BP", "MF", "CC"
  pAdjustMethod = "BH"
  #pvalueCutoff  = 0.01,
 # qvalueCutoff  = 0.05
)

# Check results
head(ego_ais_i)

```



```{r}
ego_df <- as.data.frame(ego_ais_i)

ggplot(ego_df[1:10, ], aes(x = reorder(Description, -Count), y = Count, fill = p.adjust)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_gradient(low = "red", high = "blue") +
  labs(title = "Top 10 Enriched GO Terms", x = "GO Term", y = "Gene Count") +
  theme_minimal()

```





## for mosn
```{r}
mosn.de$Gene <- rownames(mosn.de)

# Filter to only AIS genes
#ais_genes_list <- AIS_categories_df$Mouse_name
mosn.de.ais <- mosn.de[mosn.de$Gene %in% ais_genes_list, ]

#  Merge DE results with AIS metadata
annotated_mosn_ais_de <- merge(
  mosn.de.ais,
  AIS_categories_df,
  by.x = "Gene",
  by.y = "Mouse_name"
)

head(annotated_mosn_ais_de)
```





#######################################################################################new 



```{r}


# Subset upregulated in occlusion (logFC > 0)
up_occlusion_genes_m <- annotated_mosn_ais_de %>%
  filter(avg_log2FC > 0.25) %>%
  pull(Gene)

# Subset upregulated in wildtype (logFC < 0)
up_wildtype_genes_m <- annotated_mosn_ais_de %>%
  filter(avg_log2FC < -0.25) %>%
  pull(Gene)

# Background genes = all genes tested
background_genes_m <- rownames(mosn.de)

# Run enrichment
ego_occlusion_m <- enrichGO(
  gene          = up_occlusion_genes_m,
  universe      = background_genes_m,
  OrgDb         = org.Mm.eg.db,
  keyType       = "SYMBOL",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.1,
  qvalueCutoff  = 0.1
)

ego_wildtype_m <- enrichGO(
  gene          = up_wildtype_genes_m,
  universe      = background_genes_m,
  OrgDb         = org.Mm.eg.db,
  keyType       = "SYMBOL",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.1,
  qvalueCutoff  = 0.1
)


```


```{r}
# Convert results to data frames
ego_occlusion_df_m <- as.data.frame(ego_occlusion_m)
ego_wildtype_df_m <- as.data.frame(ego_wildtype_m)

# View
head(ego_occlusion_df_m)
head(ego_wildtype_df_m)

```



```{r}
ego_occlusion_df_m$Group <- "Occlusion"
ego_wildtype_df_m$Group <- "Wildtype"

combined_df_m <- rbind(ego_occlusion_df_m, ego_wildtype_df_m)

# Select top 10 from each group
top_combined_m <- combined_df_m %>%
  group_by(Group) %>%
  slice_max(order_by = Count, n = 10, with_ties = FALSE) %>%
  ungroup()

ggplot(top_combined_m, aes(x = reorder(Description, Count), y = Count, fill = Group)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  labs(title = "Top GO Terms by Condition", x = "GO Term", y = "Gene Count") +
  theme_minimal()

```

```{r}
# Top 10 enriched GO terms 
top_terms_m <- ego_occlusion_df_m[1:10, ]

ggplot(top_terms_m, aes(x = reorder(Description, p.adjust), y = p.adjust, fill = p.adjust)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_gradient(low = "blue", high = "red") +
  labs(
    title = "Top 10 Enriched GO Terms for mOSNs (Occlusion)",
    x = "GO Term",
    y = "p.adjust",
    fill = "p.adjust"
  ) +
  theme_minimal()

```


```{r}
# Top 10 enriched GO terms 
top_terms_m <- ego_occlusion_df_m[1:10, ]

ggplot(top_terms_m, aes(x = reorder(Description, pvalue), y = pvalue, fill = pvalue)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_gradient(low = "blue", high = "red") +
  labs(
    title = "Top 10 Enriched GO Terms for mOSNs (Occlusion)",
    x = "GO Term",
    y = "pvalue",
    fill = "pvalue"
  ) +
  theme_minimal()

```


```{r}
# Top 10 enriched GO terms 
top_terms_m <- ego_occlusion_df_m[1:10, ]

ggplot(top_terms_m, aes(x = reorder(Description, FoldEnrichment), y = FoldEnrichment, fill = FoldEnrichment)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_gradient(low = "blue", high = "red") +
  labs(
    title = "Top 10 Enriched GO Terms for mOSNs (Occlusion)",
    x = "GO Term",
    y = "FoldEnrichment",
    fill = "FoldEnrichment"
  ) +
  theme_minimal()

```






```{r}
# Top 10 enriched GO terms (adjust as needed)
top_terms_m <- ego_wildtype_df_m[1:10, ]

ggplot(top_terms_m, aes(x = reorder(Description, p.adjust), y = p.adjust, fill = p.adjust)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_gradient(low = "blue", high = "red") +
  labs(
    title = "Top 10 Enriched GO Terms for mOSNs (Wildtype)",
    x = "GO Term",
    y = "p.adjust",
    fill = "p.adjust"
  ) +
  theme_minimal()

```


##########


```{r}


# Extract the gene list 
gene_list_m <- annotated_mosn_ais_de$Gene

# Define background/universe genes (all genes tested in iosn.de)
background_genes_m <- rownames(mosn.de)  # 

# Run GO enrichment
ego_ais_m <- enrichGO(
  gene          = gene_list_m,
  universe      = background_genes_m,
  OrgDb         = org.Mm.eg.db,
  keyType       = "SYMBOL",
  ont           = "ALL",          # or "BP", "MF", "CC"
  pAdjustMethod = "BH"
  #pvalueCutoff  = 0.01,
 # qvalueCutoff  = 0.05
)

# Check results
head(ego_ais_m)

```


```{r}
ego_df_m <- as.data.frame(ego_ais_m)

ggplot(ego_df_m[1:10, ], aes(x = reorder(Description, -Count), y = Count, fill = p.adjust)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_gradient(low = "red", high = "blue") +
  labs(title = "Top 10 Enriched GO Terms", x = "GO Term", y = "Gene Count") +
  theme_minimal()

```


##since cannot see which is wildtype or occlusion, have to separate them out

```{r}
### for immature
# Upregulated in occlusion
i_occlusion_up_genes <- subset(iosn.de.ais, avg_log2FC > 0)$Gene

# Upregulated in wildtype
i_wildtype_up_genes <- subset(iosn.de.ais, avg_log2FC < 0)$Gene

```



```{r}
#  Run GO enrichment for occlusion-upregulated genes

background_genes_i <- rownames(iosn.de)

ego_occlusion_i <- enrichGO(
  gene          = i_occlusion_up_genes,
  universe      = background_genes_i,
  OrgDb         = org.Mm.eg.db,
  keyType       = "SYMBOL",
  ont           = "ALL",  
  pAdjustMethod = "BH"
  #pvalueCutoff  = 0.1,
  #qvalueCutoff  = 0.1
)

#  Run GO enrichment for wildtype-upregulated genes
ego_wildtype_i <- enrichGO(
  gene          = i_wildtype_up_genes,
  universe      = background_genes_i,
  OrgDb         = org.Mm.eg.db,
  keyType       = "SYMBOL",
  ont           = "ALL",
  pAdjustMethod = "BH"
  #pvalueCutoff  = 0.1,
  #qvalueCutoff  = 0.1
)

# Step 5 
head(as.data.frame(ego_occlusion_i))
head(as.data.frame(ego_wildtype_i))
```

```{r}


# Convert to data frames and add group labels
ego_occlusion_df <- as.data.frame(ego_occlusion_i) %>%
  mutate(Group = "Occlusion")

ego_wildtype_df <- as.data.frame(ego_wildtype_i) %>%
  mutate(Group = "Wildtype")

# Combine them
combined_ego_i_df <- bind_rows(ego_occlusion_df, ego_wildtype_df)

# Plot top 10 GO terms per group (adjusted if fewer than 10 terms)
combined_ego_i_df_top <- combined_ego_i_df %>%
  group_by(Group) %>%
  slice_min(order_by = p.adjust, n = 10) %>%
  ungroup()

# Barplot
ggplot(combined_ego_i_df_top, aes(x = reorder(Description, -Count), y = Count, fill = p.adjust)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_gradient(low = "red", high = "blue") +
  facet_wrap(~Group, scales = "free_y") +
  labs(
    title = "Top Enriched GO Terms in Occlusion vs Wildtype",
    x = "GO Term",
    y = "Gene Count"
  ) +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"))

```


```{r}
### for mature
# Upregulated in occlusion
m_occlusion_up_genes <- subset(mosn.de.ais, avg_log2FC > 0)$Gene

# Upregulated in wildtype
m_wildtype_up_genes <- subset(mosn.de.ais, avg_log2FC < 0)$Gene

```





```{r}
#  Run GO enrichment for occlusion-upregulated genes

background_genes_m <- rownames(mosn.de)

ego_occlusion_m <- enrichGO(
  gene          = m_occlusion_up_genes,
  universe      = background_genes_m,
  OrgDb         = org.Mm.eg.db,
  keyType       = "SYMBOL",
  ont           = "ALL",  
  pAdjustMethod = "BH"
  #pvalueCutoff  = 0.1,
  #qvalueCutoff  = 0.1
)

#  Run GO enrichment for wildtype-upregulated genes
ego_wildtype_m <- enrichGO(
  gene          = m_wildtype_up_genes,
  universe      = background_genes_m,
  OrgDb         = org.Mm.eg.db,
  keyType       = "SYMBOL",
  ont           = "ALL",
  pAdjustMethod = "BH"
  #pvalueCutoff  = 0.1,
  #qvalueCutoff  = 0.1
)

# Step 5 
head(as.data.frame(ego_occlusion_m))
head(as.data.frame(ego_wildtype_m))
```



```{r}


# Convert to data frames and add group labels
ego_occlusion_m_df <- as.data.frame(ego_occlusion_m) %>%
  mutate(Group = "Occlusion")

ego_wildtype_m_df <- as.data.frame(ego_wildtype_m) %>%
  mutate(Group = "Wildtype")

# Combine them
combined_ego_m_df <- bind_rows(ego_occlusion_m_df, ego_wildtype_m_df)

# Plot top 10 GO terms per group (adjusted if fewer than 10 terms)
combined_ego_m_df_top <- combined_ego_m_df %>%
  group_by(Group) %>%
  slice_min(order_by = p.adjust, n = 10) %>%
  ungroup()

# Barplot
ggplot(combined_ego_m_df_top, aes(x = reorder(Description, -Count), y = Count, fill = p.adjust)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_gradient(low = "red", high = "blue") +
  facet_wrap(~Group, scales = "free_y") +
  labs(
    title = "Top Enriched GO Terms in Occlusion vs Wildtype",
    x = "GO Term",
    y = "Gene Count"
  ) +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"))

```















```{r}
# Volcano plot for iOSN with AIS gene categories
ggplot(annotated_iosn_ais_de, aes(x = avg_log2FC, y = -log10(p_val_adj), color = Category)) +
  geom_point() +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "gray") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray") +
  theme_minimal() +
  labs(title = "iOSN DE Genes with AIS Annotations")

```



```{r}
# Volcano plot for iOSN with AIS gene categories
ggplot(annotated_mosn_ais_de, aes(x = avg_log2FC, y = -log10(p_val_adj), color = Category)) +
  geom_point() +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "gray") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray") +
  theme_minimal() +
  labs(title = "mOSN DE Genes with AIS Annotations")

```



```{r}
annotated_iosn_ais_de %>%
  group_by(Category) %>%
  summarise(count = n()) %>%
  ggplot(aes(x = reorder(Category, -count), y = count, fill = Category)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Number of iOSN DE Genes per AIS Category", x = "AIS Category", y = "Count")
```


```{r}
annotated_mosn_ais_de %>%
  group_by(Category) %>%
  summarise(count = n()) %>%
  ggplot(aes(x = reorder(Category, -count), y = count, fill = Category)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Number of mOSN DE Genes per AIS Category", x = "AIS Category", y = "Count")
```


```{r}
# 
DoHeatmap(harmony_nasal, features = unique(annotated_iosn_ais_de$Gene), group.by = "status") +
  ggtitle("Heatmap of DE AIS Genes in iOSN")

```


```{r}
# Example with Seurat object
DoHeatmap(harmony_nasal, features = unique(annotated_mosn_ais_de$Gene), group.by = "status") +
  ggtitle("Heatmap of DE AIS Genes in iOSN")

```




```{r}
top_ais_genes_mosn <- head(annotated_mosn_ais_de$Gene, 10)  # top 10
# Generate a list of violin plots
vplots_ais <- lapply(top_ais_genes_mosn, function(gene) {
  VlnPlot(harmony_nasal, features = gene, group.by = "status", pt.size = 0.9) +
    ggtitle(gene)
})

vplots_ais[[1]]
vplots_ais[[2]]
vplots_ais[[3]]
vplots_ais[[4]]
vplots_ais[[5]]
vplots_ais[[6]]
vplots_ais[[7]]
vplots_ais[[8]]
vplots_ais[[9]]
vplots_ais[[10]]


```



```{r}
top_ais_genes_iosn <- head(annotated_iosn_ais_de$Gene, 8)  
# Generate a list of violin plots
vplots_i_ais <- lapply(top_ais_genes_iosn, function(gene) {
  VlnPlot(harmony_nasal, features = gene, group.by = "status", pt.size = 0.9) +
    ggtitle(gene)
})

vplots_i_ais[[1]]
vplots_i_ais[[2]]
vplots_i_ais[[3]]
vplots_i_ais[[4]]
vplots_i_ais[[5]]
vplots_i_ais[[6]]
vplots_i_ais[[7]]
vplots_i_ais[[8]]


```



#Adding a new column indicating which comparison group each row belongs to
```{r}
annotated_iosn_ais_de$Region <- "iOSN"
annotated_mosn_ais_de$Region <- "mOSN"

```

#combine to one data frame
```{r}
combined_ais_de <- rbind(annotated_iosn_ais_de, annotated_mosn_ais_de)

```



#calculating -log10 pvalue
```{r}
combined_ais_de$log_pval <- -log10(combined_ais_de$p_val_adj)
combined_ais_de <- combined_ais_de[is.finite(combined_ais_de$log_pval), ]

```


```{r}
combined_ais_de$Gene <- factor(combined_ais_de$Gene, levels = unique(combined_ais_de$Gene[order(combined_ais_de$log_pval, decreasing = TRUE)]))

ggplot(combined_ais_de, aes(x = Region, y = Gene, size = log_pval, color = avg_log2FC)) +
  geom_point() +
  scale_color_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  theme_minimal() +
  labs(
    title = "-log10(p-adj) for AIS Genes: Occlusion vs Wildtype",
    x = "Region",
    y = "Gene",
    size = "-log10(p-adj)",
    color = "Log2 Fold Change"
  )
```

since avg_log2fc is calculated as avg_log2FC = log2(mean expression in occlusion / mean expression in wildtype), red means higer expression in occlusion and blue is higher in WT.













```{r}
summary(nasal_seurat_selected$TopOSN_Score1)
summary(nasal_seurat_selected$TopOSN_Score2)

```

```{r}
head(nasal_seurat_selected@meta.data[, c("TopOSN_Score1", "TopOSN_Score2")])

```









aes(x = avg_log2FC, y = -log10(p_val_adj)))


```{r}
# Replace 0 p-values with a very small number
de_results_clean <- de_results %>%
  mutate(
    p_val_adj = ifelse(p_val_adj == 0, 1e-300, p_val_adj),
    avg_log2FC = ifelse(is.infinite(avg_log2FC), NA, avg_log2FC)
  ) %>%
  filter(!is.na(avg_log2FC))

# Now plot
EnhancedVolcano(de_results_clean,
    lab = de_results_clean$gene,
    x = 'avg_log2FC',
    y = 'p_val_adj',
    title = 'Occlusion vs Wildtype',
    pCutoff = 0.05,
    FCcutoff = 1,
    pointSize = 2.5,
    labSize = 3.5
)
```





















































```{r}

obj_harmony <- readRDS("/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/obj_harmony.rds")
```

```{r}
obj_harmony = RunUMAP(obj_harmony, reduction = "harmony", dims = 1:30, reduction.name = "umap.harmony")

```


```{r}
# Color by sample or batch
DimPlot(obj_harmony, reduction = "umap.harmony", group.by = "orig.ident", label = TRUE) +
  ggtitle("UMAP Colored by Sample")

```

```{r}
FeaturePlot(
  obj_harmony,
  features = c("TopOSN_Score1", "TopOSN_Score2"),
  label = TRUE,
  reduction =  "umap.harmony"
)

```


# jus to separate by osns
```{r}
obj_harmony_3 <- obj_harmony
```



```{r}
obj_harmony_3$NeuronStage <- ifelse(
  obj_harmony_3$TopOSN_Score1 > obj_harmony_3$TopOSN_Score2, "Immature",
  ifelse(obj_harmony_3$TopOSN_Score2 > obj_harmony_3$TopOSN_Score1, "Mature", "Unassigned")
)


```


```{r}
DimPlot(obj_harmony_3, reduction = "umap.harmony", split.by = "NeuronStage", label = FALSE) +
  ggtitle("UMAP: OSN Types")

```




```{r}
#continue from here
merged_seurat_obj <- readRDS("/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/merged_seurat_obj.rds")
```


```{r}

merged_seurat_obj_copy <- merged_seurat_obj
```

```{r}
DimPlot(merged_seurat_obj, reduction = "umap", label = FALSE) +
  ggtitle("UMAP: Merged Seurat Objects")
```



```{r}
merged_seurat_obj_copy$NeuronStage <- ifelse(
  merged_seurat_obj_copy$TopOSN_Score1 > merged_seurat_obj_copy$TopOSN_Score2, "Immature",
  ifelse(merged_seurat_obj_copy$TopOSN_Score2 > merged_seurat_obj_copy$TopOSN_Score1, "Mature", "Unassigned")
)


```


```{r}
DimPlot(merged_seurat_obj_copy, reduction = "umap", split.by = "NeuronStage", label = FALSE) +
  ggtitle("UMAP: OSN Types")

```


```{r}
DimPlot(obj_harmony, reduction = "umap.harmony", group.by = "orig.ident" ,label = FALSE)
```





