---
title: "rough_code"
output: html_document
date: "2025-01-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/scratch/prj/dn_hamidlab_msc_proj/k24082190/originals")
```

```{r}
#Setting working directory
setwd("/scratch/prj/dn_hamidlab_msc_proj/k24082190/originals")
```

```{r}
#setting custom path
.libPaths("/scratch/prj/dn_hamidlab_msc_proj/Libraries")
```




```{r}
#load the libraries needed
## load dependencies
library(tidyverse)
library(Seurat)
library(SeuratObject)
library(sp)
library(scDblFinder)
library(SingleCellExperiment)
#library(SeuratWrappers)
#library(SeuratDisk)
library(patchwork)
library(clustree)
library(dplyr)
library(XML)
library(Matrix)
library(glmGamPoi)
library(gprofiler2)
library(harmony)
library(purrr)
library(limma)
library(umap)
library(pheatmap)
library(fgsea)
library(clusterProfiler)
library(org.Mm.eg.db)
library(rrvgo)
library(slingshot)
library(destiny)
library(tradeSeq)
```


#creating data
```{r}
# Create a data frame with metadata
combined_metadata <- data.frame(
  
  Path = c(
    "/scratch/prj/dn_hamidlab_msc_proj/k24082190/originals/GSE146043_RAW/wildtype1bgm/",
    "/scratch/prj/dn_hamidlab_msc_proj/k24082190/originals/GSE146043_RAW/wildtype2bgm/",
    "/scratch/prj/dn_hamidlab_msc_proj/k24082190/originals/GSE146043_RAW/wildtype3bgm/",
    "/scratch/prj/dn_hamidlab_msc_proj/k24082190/originals/GSE146043_RAW/wildtype4bgm/",
    "/scratch/prj/dn_hamidlab_msc_proj/k24082190/originals/GSE185168_RAW/mouse1/",
    "/scratch/prj/dn_hamidlab_msc_proj/k24082190/originals/GSE185168_RAW/mouse2/",
    "/scratch/prj/dn_hamidlab_msc_proj/k24082190/originals/GSE185168_RAW/mouse3/",
    "/scratch/prj/dn_hamidlab_msc_proj/k24082190/originals/GSE185168_RAW/mouse4/"
  ),
  ID = c(
    "GSE146043_1", "GSE146043_2", "GSE146043_3", "GSE146043_4",  # GSE146043 dataset
    "GSE185168_1", "GSE185168_2", "GSE185168_3", "GSE185168_4"   # GSE185168 dataset
  )
)

# View the combined metadata
print(combined_metadata)

# Save it as a CSV file
write.csv(combined_metadata, "/scratch/prj/dn_hamidlab_msc_proj/k24082190/Metadata/combined_metadata.csv", row.names = FALSE)
```


```{r}
samples.meta = read.csv("/scratch/prj/dn_hamidlab_msc_proj/k24082190/Metadata/combined_metadata.csv")

#check all paths are present
all(file.exists(samples.meta$Path))

samples.seuratlist = apply(samples.meta, 1, function(line) {
  
   
  
  # Load directory containing the matrix.mtx, features.tsv, and barcodes.tsv files provided by 10X.
  sample.data <- Read10X(line[["Path"]])
  
  
  #create seurat object
  sample.seurat = CreateSeuratObject(counts = sample.data,
                                     project = line[["ID"]])
  sample.seurat$experiment = "RNA"
  
  #adding additional metadata
  sample.seurat@meta.data = mutate(sample.seurat@meta.data, !!!line)
  
  sample.seurat
  })
  
names(samples.seuratlist) = samples.meta$ID

samples.seuratlist
```


#quantifying qc metrics
```{r}
## annotate doublets using scDblFinder
samples.seuratlist <- lapply(samples.seuratlist, function(sample){
    sce <- scDblFinder(GetAssayData(sample, assay = "RNA", layer = "counts"))
    sample$scDblFinder.class <- sce$scDblFinder.class
    sample
})
```


```{r}
## calculate percent mitochondria     
samples.seuratlist <- lapply(samples.seuratlist, function(sample){
    sample$percent_mito <- PercentageFeatureSet(sample, pattern = "^mt-",assay = "RNA") 
    Idents(sample) <- "experiment"
    sample
})
```


```{r}
samples.seuratlist[[1]]
head(samples.seuratlist[[1]]@meta.data)
```

```{r}
## plot mitochondria % distribution per experiment
do.call(wrap_plots,lapply(samples.seuratlist, function(sample){
    VlnPlot(sample, "percent_mito", group.by = "experiment", layer = "counts") +
        ggtitle(sample$orig.ident[1]) +
        scale_y_continuous(limits = c(0,65))
}))
```

```{r}
# Select the first Seurat object from the list
sample_to_plot <- samples.seuratlist[[1]]  

# Plot mitochondria percentage distribution for the selected sample
VlnPlot(sample_to_plot, "percent_mito", group.by = "experiment", assay = "RNA") +
  ggtitle(sample_to_plot$orig.ident[1]) +
  scale_y_continuous(limits = c(0, 65))

```


```{r}
## calculate percent ribosomal
samples.seuratlist <- lapply(samples.seuratlist, function(sample){
    sample$percent_ribo <- PercentageFeatureSet(sample, pattern = "^Rp[sl]",assay = "RNA") 
    Idents(sample) <- "experiment"
    sample
})
```


```{r}
## plot ribosome % distribution per experiment
do.call(wrap_plots,lapply(samples.seuratlist, function(sample){
    VlnPlot(sample, "percent_ribo", group.by = "experiment", layer = "counts") +
        ggtitle(sample$orig.ident[1]) +
        scale_y_continuous(limits = c(0,65))
}))
```

```{r}
samples.seuratlist[[1]]
head(samples.seuratlist[[1]]@meta.data)
```


```{r}
# Plotting for a single sample (GSE146043_1)
VlnPlot(samples.seuratlist$GSE146043_1, "percent_ribo", group.by = "experiment", layer = "counts") +
    ggtitle(samples.seuratlist$GSE146043_1$orig.ident[1]) +
    scale_y_continuous(limits = c(0, 65))

```


```{r}
## quantify cell cycle phase
### run cell cycle scoring
cc.genes <- cc.genes.updated.2019

#save seurat as rds file    samples.seurat list
```

```{r}
# Saving the Seurat object list (samples.seuratlist) as an RDS file
#saveRDS(samples.seuratlist, file = "/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/samples_seuratlist.rds")

samples.seuratlist <- readRDS("/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/samples_seuratlist.rds")
```




```{r}
##################
# To convert cc.genes to other organism of choice, run the following:
cc.genes <- lapply(cc.genes, function(genes) {
   gprofiler2::gorth(genes, 
        target_organism = "mmusculus"   
   )$ortholog_name
})

##################
```

```{r}

cc.genes <- lapply(cc.genes, str_to_title)

```


```{r}
head(cc.genes)
```


```{r}
samples.seuratlist <- lapply(samples.seuratlist, function(sample){
	sample <- NormalizeData(sample)
    CellCycleScoring(sample, 
                     s.features = cc.genes$s.genes, 
                     g2m.features = cc.genes$g2m.genes, 
                     set.ident = FALSE)
})
```


```{r}
do.call(wrap_plots,lapply(samples.seuratlist, function(sample){
    sample@meta.data %>% 
        ggplot(aes(x=Phase, fill = Phase)) + 
        geom_bar() +
        theme_minimal() +
        scale_fill_brewer(palette = "Set2") +
        ggtitle(sample$orig.ident[1]) 
}))
```


#Normalise, scale and regress data

```{r}
## Run SCTransform and regress out CC-score
samples.seuratlist <- lapply(samples.seuratlist, function(sample){
    SCTransform(sample, method = "glmGamPoi", 
    		vars.to.regress = c("S.Score","G2M.Score"), verbose = FALSE)
})
```


#Filtering out poor quality cells

```{r}
### check number of cells before qc
ncell_before.v <- lapply(samples.seuratlist, ncol)
ncell_before.v
```


```{r}
## 1st qc pass using median absolute deviation from the median
qc.df <- do.call(rbind,lapply(samples.seuratlist, function(sample){
  print(sample)
    feature.cutoff <- median(sample@meta.data$nFeature_SCT) - 3*mad(sample@meta.data$nFeature_SCT)
    feature.cutoff <- ifelse(feature.cutoff < 100, 100, feature.cutoff)
    mito.cutoff <- median(sample@meta.data$percent_mito) + 3*mad(sample@meta.data$percent_mito)
    mito.cutoff <- ifelse(mito.cutoff > 10, 10, mito.cutoff)
    return(data.frame(feature.cutoff = feature.cutoff,
                      mito.cutoff = mito.cutoff))
}))

qc.df[3,2] <- 25


samples.seuratlist.filt <- lapply(samples.seuratlist, function(sample) {
    sample.id <- as.character(sample$orig.ident[1])
    subset(sample, 
           subset = nFeature_SCT > qc.df[sample.id,]$feature.cutoff & percent_mito < qc.df[sample.id,]$mito.cutoff)
})

```









```{r}
## check number of cells
ncell_after.v <- lapply(samples.seuratlist.filt, ncol)
unlist(ncell_before.v) - unlist(ncell_after.v)
```


```{r}
# check metric distribution
do.call(wrap_plots,lapply(samples.seuratlist.filt, function(sample){
    VlnPlot(sample, "percent_mito", group.by = "experiment") +
        ggtitle(sample$orig.ident[1]) 
}))

do.call(wrap_plots,lapply(samples.seuratlist.filt, function(sample){
    VlnPlot(sample, "nFeature_RNA", group.by = "experiment") +
        ggtitle(sample$orig.ident[1]) 
}))
```

```{r}
# Select the specific sample 
sample_to_plot <- samples.seuratlist.filt[["GSE146043_1"]]  

# Plot percent_mito distribution for the selected sample
VlnPlot(sample_to_plot, "percent_mito", group.by = "experiment") +
    ggtitle(sample_to_plot$orig.ident[1])

```

```{r}
# Select the specific sample 
sample_to_plot <- samples.seuratlist.filt[["GSE146043_1"]]


# Plot nFeature_RNA distribution for the selected sample
VlnPlot(sample_to_plot, "nFeature_RNA", group.by = "experiment") +
    ggtitle(sample_to_plot$orig.ident[1])

```

```{r}
# overwrite original object
samples.seuratlist <- samples.seuratlist.filt
rm(samples.seuratlist.filt) 

## briefly normalise, scale, get variable features, and pca
samples.seuratlist <- lapply(samples.seuratlist, function(sample){
    SCTransform(sample, method = "glmGamPoi", verbose = FALSE)

})

```




```{r}
## 2nd qc pass: remove low quality clusters
### Briefly cluster cells using higher resolution
samples.seuratlist <- lapply(samples.seuratlist, function(sample){
	sample <- RunPCA(sample)
    sample <- FindNeighbors(sample, dims = 1:50)
    sample <- FindClusters(sample, resolution = 1.8)
    sample
})

```



```{r}
## check distribution of mito% and nFeature
do.call(wrap_plots,lapply(samples.seuratlist, function(sample){
    VlnPlot(sample, "percent_mito") +
        ggtitle(sample$orig.ident[1]) 
})) + plot_layout(ncol=1)


do.call(wrap_plots,lapply(samples.seuratlist, function(sample){
    VlnPlot(sample, "nFeature_RNA") +
        ggtitle(sample$orig.ident[1]) 
})) + plot_layout(ncol=1)
```



```{r}
## first round of removing cluster
### done by removing clusters with concomitantly low features (<1000)
### and high mito % (>2.5)
samples.seuratlist <- lapply(samples.seuratlist, function(sample){
    clusters <- sample@meta.data %>% 
        group_by(seurat_clusters) %>% 
        summarise(maxfeat = max(nFeature_RNA), 
                  maxmito = max(percent_mito)) %>% 
        filter(maxfeat >1000 | maxmito < 2.5) %>% 
        pull(seurat_clusters)
    subset(sample, idents = clusters)
})
```


```{r}
## check distribution of mito% and nFeature
do.call(wrap_plots,lapply(samples.seuratlist, function(sample){
    VlnPlot(sample, "percent_mito") +
        ggtitle(sample$orig.ident[1]) 
})) + plot_layout(ncol=1)

do.call(wrap_plots,lapply(samples.seuratlist, function(sample){
    VlnPlot(sample, "nFeature_RNA") +
        ggtitle(sample$orig.ident[1]) 
})) + plot_layout(ncol=1)
```


```{r}
## recluster and reperform qc
samples.seuratlist <- lapply(samples.seuratlist, function(sample){
    sample <- FindNeighbors(sample, dims = 1:50)
    sample <- FindClusters(sample, resolution = 1.8)
    sample
})

```



```{r}
## second round of removing cluster
### done by removing clusters with low features (<2000)
### and low rna count (<4000)
samples.seuratlist <- lapply(samples.seuratlist, function(sample){
    clusters <- sample@meta.data %>% 
        group_by(seurat_clusters) %>% 
        summarise(maxfeat = max(nFeature_RNA), 
                  maxcount = max(nCount_RNA)) %>% 
        filter(maxfeat >2000 | maxcount > 4000) %>% 
        pull(seurat_clusters)
    subset(sample, idents = clusters)
})
```



```{r}
## check distribution of nCount and nFeature
do.call(wrap_plots,lapply(samples.seuratlist, function(sample){
    VlnPlot(sample, "nCount_RNA") +
        ggtitle(sample$orig.ident[1]) 
})) + plot_layout(ncol=1)

do.call(wrap_plots,lapply(samples.seuratlist, function(sample){
    VlnPlot(sample, "nFeature_RNA") +
        ggtitle(sample$orig.ident[1]) 
})) + plot_layout(ncol=1)
```

The sctransform method models the UMI counts using a regularized negative binomial REGRESSION model to remove the variation due to sequencing depth (total nUMIs per cell), while adjusting the variance based on pooling information across genes with similar abundances
```{r}
## Final SCTransform
samples.seuratlist <- lapply(samples.seuratlist, function(sample){
    SCTransform(sample, method = "glmGamPoi", 
    		vars.to.regress = c("S.Score","G2M.Score"), verbose = FALSE)
})
```
#New RDS file
```{r}
#saveRDS(samples.seuratlist, file = "/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/FinalSCT_samples_seuratlist.rds")

### from here load in this rds and continue::::::::
#samples.seuratlist <- readRDS("/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/FinalSCT_samples_seuratlist.rds")
```




```{r}
## Perform UMAP reduction
samples.seuratlist <- lapply(samples.seuratlist, function(sample){
	sample <- RunPCA(sample)
	sample <- RunTSNE(sample, dims = 1:50)
    sample <- RunUMAP(sample, dims = 1:50)
    sample 
})
```


```{r}
#saveRDS(samples.seuratlist, file = "/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/UMAP_Reduction_samples_seuratlist.rds")

### from here load in this rds and continue::::::::
samples.seuratlist <- readRDS("/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/UMAP_Reduction_samples_seuratlist.rds")
```




#cell clustering

```{r}
## Determine the best clustering resolution
### Set range of resolutions
res.range <- seq(0.4, 3, 0.2)

samples.seuratlist <- lapply(samples.seuratlist, function(obj){
    DefaultAssay(obj) <- "SCT"
    obj <- FindNeighbors(obj, dims = 1:50)
    for(res in res.range){
        obj <- FindClusters(obj, resolution = res)
    }
    obj
})
```


```{r}
# plot clustree for each sample
lapply(samples.seuratlist, function(dat){
    clustree::clustree(dat, prefix = "SCT_snn_res.")   # change the prefix accordingly
})  
```


```{r}
### Calculate weighted sum of squares for each experiment
wss.all <- do.call(bind_rows,lapply(samples.seuratlist, function(sample){
  wss.out <- do.call(bind_rows, lapply(res.range, function(k){

    # get centroid per cluster
    centroids <- aggregate(Embeddings(sample, reduction = "pca"),
          sample[[paste0("SCT_snn_res.", k)]], 
          mean) %>% 
    column_to_rownames(paste0("SCT_snn_res.", k))
    diffsq <- (Embeddings(sample, reduction = "pca") - 
        centroids[sample@meta.data[[paste0("SCT_snn_res.", k)]],])^2
    sdsq <- colSds(Embeddings(sample, reduction = "pca"))^2
    wdiff <- diffsq/sdsq
    
    wss <- sum(sqrt(rowSums(wdiff)))
    data.frame(clust=k, wss = wss)
 }))
  
  wss.out %>% 
    mutate(id = sample$orig.ident[1])
}))
```


```{r}
### Plot out WSS vs resolution
wss.all %>% 
  group_by(id) %>% 
  mutate(scaled_wss = scale(wss)) %>% 
  ggplot(aes(x=clust,y=scaled_wss,group=id, colour=id)) +
  geom_line()
```

```{r}
### Cluster for real
res <- 1 
samples.seuratlist <- lapply(samples.seuratlist, function(sample){
	sample <- FindNeighbors(sample, verbose = F)
    sample <- FindClusters(sample, resolution = res, verbose = F)
})
```


```{r}
##rds file after clustering
#saveRDS(samples.seuratlist, file = "/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/After_Clustering_samples_seuratlist.rds")

#continue from here
samples.seuratlist <- readRDS("/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/After_Clustering_samples_seuratlist.rds")
```




```{r}
lapply(samples.seuratlist, DimPlot, label= TRUE)
```

```{r}
#VariableFeatures(samples.seuratlist$GSE146043_1)


DotPlot(samples.seuratlist$GSE146043_1, features = c("Syn1" ), split.by = NULL)

```


```{r}
FeaturePlot(
  samples.seuratlist$GSE146043_1,
  features = c("Syn1") 
  
)

```




```{r}
#importing mouse genes
mouse_genes = read.csv("/scratch/prj/dn_hamidlab_msc_proj/k24082190/originals/Dishalas_Filtered_OSN_Genes.csv")
```

```{r}
#importing human genes
human_genes = read.csv("/scratch/prj/dn_hamidlab_msc_proj/k24082190/originals/Filtered_human_genes.csv")
```


```{r}
# Filter the data for mOSN (mature OSN) cluster
mouse_mOSN_info = mouse_genes[mouse_genes$cluster == "mOSN", ]

# Filter the data for iOSN (immature OSN) cluster
mouse_iOSN_info = mouse_genes[mouse_genes$cluster == "iOSN", ]
```

```{r}
human_gene_lists = split(human_genes$gene, human_genes$cluster)

# Separate into iOSN and mOSN genes
human_mOSN_genes = human_genes[human_genes$cluster =="Mature Neurons", ]
human_iOSN_genes = human_genes[human_genes$cluster =="Immature Neurons", ]


# Convert human mOSN genes to mouse orthologs
converted_human_mOSN_genes = gprofiler2::gorth(
  human_mOSN_genes$gene, 
  source_organism = "hsapiens", 
  target_organism = "mmusculus"
)

# Convert human iOSN genes to mouse orthologs
converted_human_iOSN_genes = gprofiler2::gorth(
  human_iOSN_genes$gene, 
  source_organism = "hsapiens", 
  target_organism = "mmusculus"
) 
```


```{r}
# Extract only necessary columns
converted_human_mOSN_df <- converted_human_mOSN_genes[, c("input", "ortholog_name")]
converted_human_iOSN_df <- converted_human_iOSN_genes[, c("input", "ortholog_name")]

# Renaming the column nmes
colnames(converted_human_mOSN_df) <- c("gene", "mouse_gene")
colnames(converted_human_iOSN_df) <- c("gene", "mouse_gene")

# Merge back with original data to preserve other columns like p-value, avg_logFC and so on
human_mOSN_info <- merge(human_mOSN_genes, converted_human_mOSN_df, 
                         by = "gene", all.x = TRUE)

human_iOSN_info <- merge(human_iOSN_genes, converted_human_iOSN_df, 
                         by = "gene", all.x = TRUE)

# Add source column
human_mOSN_info$source <- "converted_human_mOSN"
human_iOSN_info$source <- "converted_human_iOSN"

# View results
head(human_mOSN_info)
head(human_iOSN_info)





```

## need to combine dataframe

```{r}


# Standardisinge column names by ensuring all data frames have the same structure
human_iOSN_info$species <- "Human"
human_mOSN_info$species <- "Human"
mouse_mOSN_info$species <- "Mouse"
mouse_iOSN_info$species <- "Mouse"

# Add missing columns (source in mouse dataset)
mouse_mOSN_info$source <- "converted_mouse_mOSN"
mouse_iOSN_info$source <- "converted_mouse_iOSN"

# Making sure all have the same column order
all_columns <- c("gene", "p_val", "avg_logFC", "pct.1", "pct.2", "p_val_adj", "cluster", "mouse_gene", "source", "species")

# Add missing columns to mouse data frames
mouse_mOSN_info$mouse_gene <- mouse_mOSN_info$gene
mouse_iOSN_info$mouse_gene <- mouse_iOSN_info$gene

# Bind all data frames
combined_df <- bind_rows(
  dplyr::select(mouse_mOSN_info, all_of(all_columns)),  
  dplyr::select(mouse_iOSN_info, all_of(all_columns)),
  dplyr::select(human_iOSN_info, all_of(all_columns)),  
  dplyr::select(human_mOSN_info, all_of(all_columns))
)

# View the combined data frame
head(combined_df)



```


```{r}
combined_df %>%
  distinct(mouse_gene, .keep_all = T) %>%   ##this line gets the distinct genes, gets rid of human duplicate
  pull(cluster) %>% unique()
```



```{r}


# Get list of genes for each mouse sample 
mouse_gene_lists <- lapply(samples.seuratlist, function(sample) {
    rownames(sample)  # Extract gene names from each Seurat object
})


# Find genes common in all mouse samples
common_mouse_genes <- Reduce(intersect, mouse_gene_lists)

# Filter the dataframe: Keep human genes & only common mouse genes
filtered_df <- combined_df %>%
  filter(mouse_gene %in% common_mouse_genes)

# View the filtered data
head(filtered_df)

```

```{r}
# Find genes that are in combined_df but not in filtered_df
removed_genes <- setdiff(combined_df$mouse_gene, filtered_df$mouse_gene)

# View the removed genes
removed_genes                        ####to see which genes are not common amongst the 8 samples, includes human genes

```

```{r}
# Check if removed genes are from Human or Mouse
removed_genes_info <- combined_df %>%
  filter(gene %in% removed_genes) %>%
  dplyr::select(gene, species, source)

# View the removed genes' details
removed_genes_info

```

```{r}
#just checking where one of these genes is missing

str_subset(rownames(samples.seuratlist$GSE146043_3), "Fam213b")

 
str_subset(rownames(samples.seuratlist$GSE146043_4), "Fam213b")


str_subset(rownames(samples.seuratlist$GSE185168_1), "Fam213b")

##missing in GSE185168_1
```

```{r}
filtered_df %>%
  distinct(mouse_gene, .keep_all = T) %>%   ##this line gets the distinct genes, gets rid of human duplicate
  pull(cluster) %>% unique()
```

```{r}
# Recode human clusters to match mouse clusters
filtered_df <- filtered_df %>%
  mutate(cluster = case_when(
    cluster == "Immature Neurons" ~ "iOSN",  # Recode "Immature Neurons" to "iOSN"
    cluster == "Mature Neurons" ~ "mOSN",    # Recode "Mature Neurons" to "mOSN"
    TRUE ~ cluster                          # Keep other clusters unchanged
  ))
```


```{r}


# getting the top mOSN 
#top_100_mOSN <- filtered_df %>%
 # filter(cluster == "mOSN") %>%   # Filter for mOSN cluster
 # arrange(desc(avg_logFC)) %>%  # Sort avg_logFC in descending order
  #head(100)     # Select top 100 genes

# getting the top iOSN
#top_100_iOSN <- filtered_df %>%
  #filter(cluster == "iOSN") %>%  # Filter for iOSN cluster
  #arrange(desc(avg_logFC)) %>%   # Sort avg_logFC in descending order
  #head(100)   # Select top 100 genes


filtered_df %>%
  group_by(cluster) %>%
  slice_max(n=100, order_by = avg_logFC) %>%  ##getting the top 100 for mOSN and iOSN each according to avg_logFC
  group_by(species, cluster) %>%
  tally()
```



```{r}

##checking to see how many mice and human genes are in my two top 100 genes

# For mOSN top 100
#mouse_human_count_mOSN <- table(top_100_mOSN$species)
#print(mouse_human_count_mOSN)

# For iOSN top 100
#mouse_human_count_iOSN <- table(top_100_iOSN$species)
#print(mouse_human_count_iOSN)

```

```{r}
# Check for duplicate rows
#duplicates <- filtered_df[duplicated(filtered_df), ]

# View the duplicated rows
#print(duplicates)

```


```{r}
# Check for duplicate rows in the mOSN top 100
#duplicates_mOSN <- top_100_mOSN[duplicated(top_100_mOSN), ]

# View the duplicated rows in mOSN top 100
#print(duplicates_mOSN)

```

```{r}
# Check for duplicate rows in the iOSN top 100
#duplicates_iOSN <- top_100_iOSN[duplicated(top_100_iOSN), ]

# View the duplicated rows in iOSN top 100
#print(duplicates_iOSN)

```

#extracting iOSN and mOSN genes separately
```{r}
top_mOSN_genes <- filtered_df %>%
  filter(cluster == "mOSN") %>%
  slice_max(n = 100, order_by = avg_logFC) %>%
  pull(gene)

top_iOSN_genes <- filtered_df %>%
  filter(cluster == "iOSN") %>%
  slice_max(n = 100, order_by = avg_logFC) %>%
  pull(gene)

# Store in a list for AddModuleScore
top_osn_gene_sets <- list("iOSN" = top_iOSN_genes, "mOSN" = top_mOSN_genes)

```









#doing the add module score
```{r}
samples.seuratlist <- lapply(samples.seuratlist, function(sample) {
  AddModuleScore(
    object = sample,
    features = top_osn_gene_sets,  # List of gene sets
    name = "TopOSN_Score",  
    ctrl = 100, 
    assay = "RNA"  
  )
})



```


```{r}
FeaturePlot(
  samples.seuratlist$GSE146043_1, label = TRUE,
  features = c("TopOSN_Score1","TopOSN_Score2") 
  
)
```

###checking the top 1 and top 2 OSN scores for one sample
```{r}
samples.seuratlist$GSE185168_1@meta.data
```


```{r}


# Set cluster identities to "SCT_snn_res.0.8"
Idents(samples.seuratlist$GSE146043_2) <- "SCT_snn_res.0.8"

# Feature plots for TopOSN_Score1 and TopOSN_Score2 for sample 2
FeaturePlot(samples.seuratlist$GSE146043_2, 
            features = "TopOSN_Score1", 
            label = TRUE, 
            min.cutoff = 0)

FeaturePlot(samples.seuratlist$GSE146043_2, 
            features = "TopOSN_Score2", 
            label = TRUE, 
            min.cutoff = 0)

# Summarise TopOSN scores per cluster
samples.seuratlist$GSE146043_2@meta.data %>%
  group_by(SCT_snn_res.0.8) %>%
  summarise(
    TopOSN_Score1_avg = mean(TopOSN_Score1, na.rm = TRUE),
    TopOSN_Score1_exp_perc = sum(TopOSN_Score1 > 0.1, na.rm = TRUE) / n(),
    TopOSN_Score2_avg = mean(TopOSN_Score2, na.rm = TRUE),
    TopOSN_Score2_exp_perc = sum(TopOSN_Score2 > 0.1, na.rm = TRUE) / n()
  )





```


```{r}
cluster_OSN_avg_scores <- lapply(samples.seuratlist, function(sample) {
  sample@meta.data %>%
    group_by(SCT_snn_res.0.8) %>%
    summarise(
      TopOSN_Score1_avg = mean(TopOSN_Score1, na.rm = TRUE),
      TopOSN_Score1_exp_perc = sum(TopOSN_Score1 > 0.1, na.rm = TRUE) / n(),
      TopOSN_Score2_avg = mean(TopOSN_Score2, na.rm = TRUE),
      TopOSN_Score2_exp_perc = sum(TopOSN_Score2 > 0.1, na.rm = TRUE) / n()
    )
})

# Assign names to the list to keep track of each dataset
names(cluster_OSN_avg_scores) <- names(samples.seuratlist)


cluster_OSN_avg_scores




```


```{r}

lapply(names(samples.seuratlist), function(sample) {
  Idents(samples.seuratlist[[sample]]) <- "SCT_snn_res.0.8"  #### Set cluster identities based on SCT_snn_res.0.8
  
  # Generate FeaturePlots
  Score1plot <- FeaturePlot(samples.seuratlist[[sample]], features = "TopOSN_Score1", 
                            min.cutoff = 0, label = TRUE) + 
        ggtitle(paste(sample, "- TopOSN_Score1"))
  
  Score2plot <- FeaturePlot(samples.seuratlist[[sample]], features = "TopOSN_Score2", 
                            min.cutoff = 0, label = TRUE) + 
        ggtitle(paste(sample, "- TopOSN_Score2"))

  # Print the plots
  print(Score1plot)
  print(Score2plot)
})




```


```{r}


# Extract the first sample
first_sample <- names(samples.seuratlist)[1]
Idents(samples.seuratlist[[first_sample]]) <- "SCT_snn_res.0.8"

# Generate the two FeaturePlots
Score1plot <- FeaturePlot(samples.seuratlist[[first_sample]], features = "TopOSN_Score1", 
                          min.cutoff = 0, label = TRUE) + 
      ggtitle(paste(first_sample, "- TopOSN_Score1"))

Score2plot <- FeaturePlot(samples.seuratlist[[first_sample]], features = "TopOSN_Score2", 
                          min.cutoff = 0, label = TRUE) + 
      ggtitle(paste(first_sample, "- TopOSN_Score2"))

# Plot side by side
Score1plot + Score2plot

```




```{r}
# plot clustree for one sample
clustree::clustree(samples.seuratlist$GSE146043_1, prefix = "SCT_snn_res.")   


  
```

```{r}
# Choosing clusters wanted
selected_clusters <- list(
  GSE146043_1 = c("0","1","3", "9"),  
  GSE146043_2 = c("1", "2","3","7","8"),
  GSE146043_3 = c("0", "1", "2","3","4"),
  GSE146043_4 = c("0","1","2", "7"),
  GSE185168_1 = c("0","1", "2","3", "4","5","6","7","8"),
  GSE185168_2 = c("0", "1","2","3","4","7"),
  GSE185168_3 = c("0", "1","3","4","5","6","9"),
  GSE185168_4 = c("0", "1","4")
)

```



```{r}
# Filter each Seurat object based on selected clusters
filtered_seurat_objects <- lapply(names(samples.seuratlist), function(sample) {
  selected_clusters <- filtered_cluster_OSN_avg_scores[[sample]]$SCT_snn_res.0.8
  
  subset(samples.seuratlist[[sample]], SCT_snn_res.0.8 %in% selected_clusters)
})

# Assign names to keep track
names(filtered_seurat_objects) <- names(samples.seuratlist)

head(filtered_seurat_objects[[1]]@meta.data)  # Check first few rows of metadata
filtered_seurat_objects[[1]]  # basic info about the first Seurat object



```

```{r}
lapply(filtered_seurat_objects, function(obj) unique(obj$SCT_snn_res.0.8)) ##listing the clusters

```





```{r}

merged_seurat_obj <- merge(filtered_seurat_objects[[1]],  ##first object
                           y = filtered_seurat_objects[-1],  #everything but the first object
                           add.cell.ids = names(filtered_seurat_objects))

# View merged object
print(merged_seurat_obj)


```


```{r}
head(merged_seurat_obj@meta.data) # View metadata
```


#Running SCT transform
```{r}
merged_seurat_obj <- SCTransform(merged_seurat_obj, method = "glmGamPoi", verbose = FALSE)
```


```{r}
merged_seurat_obj <- RunPCA(merged_seurat_obj, verbose = FALSE)

# Find neighbors based on PCA
merged_seurat_obj <- FindNeighbors(merged_seurat_obj, dims = 1:50)

# Find clusters 
merged_seurat_obj <- FindClusters(merged_seurat_obj, resolution = 2)

# Visualise clusters using UMAP
merged_seurat_obj <- RunUMAP(merged_seurat_obj, dims = 1:50)
DimPlot(merged_seurat_obj, reduction = "umap", label = TRUE, group.by = "orig.ident")

```


```{r}
#saveRDS(merged_seurat_obj, file = "/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/merged_seurat_obj.rds")

#continue from here
merged_seurat_obj <- readRDS("/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/merged_seurat_obj.rds")
```

From the umap above we can see quite strong batch effects from the two experiments

#running harmony with orig.ident as batch variable
```{r}
merged_seurat_obj <- RunHarmony(merged_seurat_obj, group.by.vars = "orig.ident", assay.use = "SCT")
```


```{r}
# Run PCA on the Harmony-corrected data
merged_seurat_obj <- RunPCA(merged_seurat_obj, assay = "SCT", reduction.key = "harmony_")

```


```{r}
# Run UMAP on PCA dimensions
merged_seurat_obj <- RunUMAP(merged_seurat_obj, dims = 1:30)

# Find clusters 
merged_seurat_obj <- FindClusters(merged_seurat_obj, resolution = 1.8)

```

```{r}
# Visualise Harmony-corrected UMAP
DimPlot(merged_seurat_obj, reduction = "umap", label = TRUE, group.by = "orig.ident")

```


```{r}
# Find clusters
merged_seurat_obj <- FindClusters(merged_seurat_obj, resolution = 1.8)

# Visualise clusters on UMAP
DimPlot(merged_seurat_obj, reduction = "umap", label = TRUE)
merged_seurat_obj@meta.data$label <-
  ifelse(merged_seurat_obj$seurat_clusters %in% iOSN, "iOSN",
         ifelse(merged_seurat_obj$seurat_clusters %in% mOSN, "mOSN", "NA"))

```



#Running rPCA
```{r}
options(future.globals.maxSize = +Inf)

obj <- IntegrateLayers(
  object = merged_seurat_obj, method = RPCAIntegration, normalization.method = "SCT",
  orig.reduction = "pca", new.reduction = "integrated.rpca",
  verbose = FALSE
)


```

```{r}
obj <- RunUMAP(obj, reduction = "integrated.rpca", dims = 1:30, reduction.name = "umap.rPCA")
p1 <- DimPlot(
  obj,
  reduction = "umap.rPCA",
  group.by = c("orig.ident")
  #split.by = "orig.ident"
)

obj <- FindNeighbors(obj, reduction = "integrated.rpca", dims = 1:30)
obj <- FindClusters(obj, resolution = 2, cluster.name = "rPCA_clusters")
```

```{r}
p1
```


## doing cca
```{r}
obj_cca <- IntegrateLayers(
  object = merged_seurat_obj, method = CCAIntegration, normalization.method = "SCT",
  orig.reduction = "pca", new.reduction = "integrated.cca",
  verbose = FALSE
)
```


```{r}
obj_cca <- RunUMAP(obj_cca, reduction = "integrated.cca", dims = 1:30, reduction.name = "umap.cca")
DimPlot(
  obj_cca,
  reduction = "umap.cca",
  group.by = c("orig.ident")
 # split.by = "orig.ident"
  
)

obj_cca <- FindNeighbors(obj_cca, reduction = "integrated.cca", dims = 1:30)
obj_cca <- FindClusters(obj_cca, resolution = 2, cluster.name = "scvi_clusters")
```

```{r}
obj_cca
```





# just checking for this gene
```{r}
FeaturePlot(
  samples.seuratlist$GSE146043_1,
  features = c("Tubb3"),
  label = TRUE
  
)
```

#harmony again
```{r}
obj_harmony <- IntegrateLayers(
  object = merged_seurat_obj, method = HarmonyIntegration, normalization.method = "SCT",
  orig.reduction = "pca", new.reduction = "harmony",
  verbose = FALSE
)
```



```{r}
obj_harmony <- RunUMAP(obj_harmony, reduction = "harmony", dims = 1:30, reduction.name = "umap.harmony")
p1_harmony <- DimPlot(
  obj_harmony,
  reduction = "umap.harmony",
  group.by = c("harmony_clusters")
  #split.by = "orig.ident"
  
)

 FeaturePlot(
  obj_harmony,
  features = c("TopOSN_Score1", "TopOSN_Score2"),
  label = TRUE,
  reduction =  "umap.harmony"
  
)
head(obj_harmony)

obj_harmony <- FindNeighbors(obj_harmony, reduction = "harmony", dims = 1:30)
obj_harmony <- FindClusters(obj_harmony, resolution = 1, cluster.name = "harmony_clusters")
```


```{r}
obj_harmony
```


```{r}
p1_harmony
```

```{r}
obj_harmony <- FindClusters(obj_harmony, resolution = 0.5, cluster.name = "harmony_clusters")

DimPlot(obj_harmony, group.by = "harmony_clusters", reduction = "umap.harmony")

# Idents(merged_seurat_obj) <- "seurat_clusters"  # Ensure correct clustering identity
# table(Idents(merged_seurat_obj))  # Check cluster distribution

```

```{r}
merged_cluster_OSN_avg_scores <- obj_harmony@meta.data %>%
  group_by(harmony_clusters) %>%
  summarise(
    TopOSN_Score1_avg = mean(TopOSN_Score1, na.rm = TRUE),
    TopOSN_Score1_exp_perc = sum(TopOSN_Score1 > 0.1, na.rm = TRUE) / n(),
    TopOSN_Score2_avg = mean(TopOSN_Score2, na.rm = TRUE),
    TopOSN_Score2_exp_perc = sum(TopOSN_Score2 > 0.1, na.rm = TRUE) / n()
  )

# View the result
print(merged_cluster_OSN_avg_scores)

```




#creating list of clusters that need to be removed
```{r}
clusters_to_keep = as.character(c(10,13,17,0:9,15))
```


```{r}
obj_harmony <- subset(obj_harmony, idents = clusters_to_keep)
```



```{r}
obj_harmony = SCTransform(obj_harmony, method = "glmGamPoi", assay = "RNA",
                               vars.to.regress = c("S.Score", "G2M.Score"), verbose = FALSE)
```




```{r}
# PCA first
obj_harmony <- RunPCA(obj_harmony, verbose = FALSE)

obj_harmony=
  IntegrateLayers(
  object = obj_harmony, method = HarmonyIntegration, normalization.method = "SCT",
  orig.reduction = "pca", new.reduction = "harmony",
  verbose = FALSE
)
```





```{r}
obj_harmony = RunUMAP(obj_harmony, reduction = "harmony", dims = 1:30, reduction.name = "umap.harmony")
DimPlot(
  obj_harmony,
  reduction = "umap.harmony",
  group.by = c("orig.ident"),
  split.by = "orig.ident",
  ncol=2
  
)

 FeaturePlot(
  obj_harmony,
  features = c("TopOSN_Score1, TopOSN_Score2"),
  label = TRUE,
  reduction =  "umap.harmony"
  
)


# FindNeighbors(obj_harmony, reduction = "harmony", dims = 1:30)
# FindClusters(obj_harmony, resolution = 2, cluster.name = "harmony_clusters")
```

```{r}
#saveRDS(obj_harmony, file = "/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/obj_harmony.rds")

#continue from here
#obj_harmony <- readRDS("/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/obj_harmony.rds")
```

```{r}
FeaturePlot(
  obj_harmony,
  features = c("TopOSN_Score1", "TopOSN_Score2"),
  label = TRUE,
  reduction =  "umap.harmony"
)

```

#read in the mice AIS gene list
```{r}
mice_AIS_genes = read.csv("/scratch/prj/dn_hamidlab_msc_proj/k24082190/originals/AIS_list.csv")
```

#creating list of AIS genes from the mice list
```{r}
mice_AIS_gene_list = mice_AIS_genes$Symbol
```

#reading in the human AIS gene list    
```{r}
AIS_human_genes = read.csv("/scratch/prj/dn_hamidlab_msc_proj/k24082190/originals/AIS_human_genes.csv", fileEncoding = "Latin1")

## encoding error so used latin1
```

#making list of human AIS genes
```{r}
AIS_human_gene_list = AIS_human_genes$Symbol
```





```{r}
AIS_mouse_orthologues <- sapply(AIS_human_gene_list, function(gene) {
  result <- gprofiler2::gorth(gene, target_organism = "mmusculus")
  
  if (!is.null(result) && nrow(result) > 0) {     ## Check if the result is not NULL and contains at least one row
    return(result$ortholog_name[1])  # Select the first orthologue only
  } else {
    return(NA)  # If no orthologue, return NA
  }
})

# Convert into a DataFrame
AIS_orthology_df <- data.frame(
  Human_Gene = AIS_human_gene_list,
  Mouse_Orthologue = AIS_mouse_orthologues
)

# View the result
View(AIS_orthology_df)
print(AIS_orthology_df)

# '&&' to check if both conditions are true
# I used 'sapply' instead of 'lapply' because it would create columns for some individual genes 
```


#combining them and removing duplicates  179 in total

```{r}
# Combine both lists
combined_mice_AIS_genes <- unique(c(mice_AIS_gene_list, AIS_orthology_df$Mouse_Orthologue))

# Remove any NA values if present
combined_mice_AIS_genes <- na.omit(combined_mice_AIS_genes)

# View the combined list
print(combined_mice_AIS_genes)
```

# overiding previous rds file
```{r}
#saveRDS(obj_harmony, file = "/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/obj_harmony.rds")

#continue from here
obj_harmony <- readRDS("/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/obj_harmony.rds")
```



# Loading in the neocortical seurat that was the h5ad file

```{r}
load("/scratch/prj/dn_hamidlab_msc_proj//k24082190/Seurat.ss.rda")
```

```{r}
cc.genes <- cc.genes.updated.2019

```

```{r}
# To convert cc.genes to other organism of choice
cc.genes <- lapply(cc.genes, function(genes) {
   gprofiler2::gorth(genes, 
        target_organism = "mmusculus"   #change target_organism accordingly
   )$ortholog_name
})

```



#setting assay to RNA explicly
```{r}
DefaultAssay(ss.seurat) <- "RNA"

```

#checking for na's
```{r}
sum(is.na(cc.genes$s.genes))  # Check for NA in s.genes
sum(is.na(cc.genes$g2m.genes))  # Check for NA in g2m.genes

```




# it seems the problem stems from the fact that the seurat object uses ID's and cc.genes uses symbols
```{r}
# Check the first few gene names in cc.genes
head(cc.genes$s.genes)
head(cc.genes$g2m.genes)

# Check the first few row names in ss.seurat
tail(rownames(ss.seurat))

```










```{r}
ss.seurat <- CellCycleScoring(ss.seurat, 
                 s.features = cc.genes$s.genes, 
                 g2m.features = cc.genes$g2m.genes, 
                 set.ident = FALSE)


```

```{r}
head(ss.seurat@meta.data)
```

#subsetting ss.seurat so we only include the visual cortex
```{r}
ss.seurat <- subset(ss.seurat, subset = region_label == "VISp")

```

#removing cells that are non neuronal, only keeping those in 'class label' that are either Glutamatergic"or "GABAergic

```{r}
ss.seurat <- subset(ss.seurat, subset = class_label %in% c("Glutamatergic", "GABAergic"))

```


```{r}

options(future.globals.maxSize = 18000 * 1024^2)      
ss.seurat <- SCTransform(ss.seurat, method = "glmGamPoi", 
    		vars.to.regress = c("S.Score","G2M.Score"), verbose = FALSE)
```

```{r}
#saveRDS(ss.seurat, file = "/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/ss_seurat.rds")

#continue from here
ss.seurat <- readRDS("/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/ss_seurat.rds")
```


```{r}
#running PCA and UMAP
ss.seurat <- RunPCA(ss.seurat, verbose = FALSE)
ss.seurat <- RunUMAP(ss.seurat, dims = 1:50, verbose = FALSE)

```

```{r}
DimPlot(ss.seurat, reduction = "umap", label = TRUE, group.by = "class_label" )
```

```{r}
DimPlot(ss.seurat, reduction = "umap", label = TRUE, group.by = "subclass_label" )
```



```{r}
head(ss.seurat)
```


```{r}
head(obj_harmony)
```

## trying to fix the 'Error in .subscript.2ary(x, , j, drop = drop) : subscript out of bounds' when merging
```{r}

obj_harmony@meta.data[is.na(obj_harmony@meta.data)] <- "NA"
ss.seurat@meta.data[is.na(ss.seurat@meta.data)] <- "NA"

```

## still trying to fix error mentioned above, 
```{r}
DefaultAssay(obj_harmony) <- "SCT"
DefaultAssay(ss.seurat) <- "SCT"

```

##checking the genes
```{r}
AIS_harmony <- setdiff(rownames(combined_mice_AIS_genes), rownames(obj_harmony[["SCT"]]@data))

```

```{r}
AIS_harmony
```

```{r}
setdiff(rownames(combined_mice_AIS_genes), rownames(ss.seurat[["SCT"]]@data))
```



#merging the two seurat objects
```{r}
# Classifying cells in obj_harmony based on TopOSN_Score1 and TopOSN_Score2
obj_harmony$cell_type <- ifelse(obj_harmony$TopOSN_Score1 > obj_harmony$TopOSN_Score2, "Immature Neuron", 
                         ifelse(obj_harmony$TopOSN_Score2 > obj_harmony$TopOSN_Score1, "Mature Neuron", "Undefined"))

# assign cell type labels from ss.seurat
ss.seurat$cell_type <- ss.seurat$class_label  # "Glutamatergic" or "GABAergic"

#add dataset labels
obj_harmony$dataset <- "obj_harmony"
ss.seurat$dataset <- "ss_seurat"

# Change default assay to RNA and drop SCT assay from both Seurat objects like the dude from github said
DefaultAssay(obj_harmony) <- "RNA"
DefaultAssay(ss.seurat) <- "RNA"
obj_harmony[['SCT']] <- NULL
ss.seurat[['SCT']] <- NULL

# Merge Seurat objects
fused_seurat <- merge(x = obj_harmony, y = list(ss.seurat))






```


solution to the error above was found at https://github.com/satijalab/seurat/issues/8561





```{r}
# Rerun SCTransform  
fused_seurat <- SCTransform(fused_seurat, verbose = FALSE)

#Run PCA and UMAP for visualization
fused_seurat <- RunPCA(fused_seurat, verbose = FALSE)
fused_seurat <- RunUMAP(fused_seurat, dims = 1:30)

# Plot UMAP to check if immature and mature neurons separate well
DimPlot(fused_seurat, group.by = "cell_type", label = TRUE) + NoLegend()

```



## saving RDS file for fused_seurat
```{r}
#saveRDS(fused_seurat, file = "/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/fused_seurat.rds")

#continue from here
fused_seurat <- readRDS("/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/fused_seurat.rds")
```

## following first pipeline. On the sct slot in merged object got to data, get top 2000 variable genes, scale and then do PCA.

# getting top 2000 variable genes from sct in data

```{r}
thelogcounts <- as.matrix(fused_seurat[["SCT"]]@data)

geneVars <- rowVars(thelogcounts)

```

```{r}
geneVars <- sort(geneVars, decreasing = TRUE)
```

```{r}
top_2000_geneVars <- (geneVars [1:2000] )  
```


## doing prcomp

```{r}
pca_uncorrected<-prcomp(t(thelogcounts[names(top_2000_geneVars),]), center = TRUE, scale. = TRUE)
```

```{r}
pca_uncorrected$x[1:5, 1:5]
```

```{r}
#plotting
grp <- fused_seurat$orig.ident[fused_seurat$orig.ident != "CTX-HPF_smartseq"]
pca_uncorrected$x %>%
  as.data.frame() %>% 
  # filter(PC1 < -25) %>%
ggplot(aes(x = PC1, y = PC2, color = fused_seurat$orig.ident)) +
  geom_point() +  # Adds points
  labs(title = "PCA: PC1 vs PC2", x = "PC1", y = "PC2") +
  theme_minimal()
```

##starting new pipeline . Removing batch effect from merged object, get top 2000 variable genes, scale and prcomp



```{r}
thelogcounts_2 <- as.matrix(fused_seurat[["SCT"]]@data)

```

```{r}
# Remove batch effect
model.matrix(~fused_seurat$cell_type)
logcounts_matrix_rbe2 <- limma::removeBatchEffect(thelogcounts_2, batch = fused_seurat$orig.ident, design = model.matrix(~fused_seurat$cell_type))

```


















```{r}
head(thelogcounts_2) [1:5, 1:5]
```

# when you come back to this remember to replace them with logcounts_matrix_rbe2
```{r}
geneVars_2 <- rowVars(logcounts_matrix_rbe)
```

```{r}
geneVars_2 <- sort(geneVars_2, decreasing = TRUE)
```


```{r}
top_2000_geneVars_2 <- (geneVars_2 [1:2000] )
```

```{r}
pca_corrected<-prcomp(t(thelogcounts_2[names(top_2000_geneVars_2),]), center = TRUE, scale. = TRUE)
```

```{r}
pca_corrected$x[1:5, 1:5]
```

```{r}
#plotting
grp_2 <- fused_seurat$orig.ident[fused_seurat$orig.ident != "CTX-HPF_smartseq"]
pca_corrected$x %>%
  as.data.frame() %>% 
  #filter(PC1 < -25) %>%
ggplot(aes(x = PC1, y = PC2, color = fused_seurat$orig.ident)) +
  geom_point() +  # Adds points
  labs(title = "PCA: PC1 vs PC2", x = "PC1", y = "PC2") +
  theme_minimal()
```


#Get UMAP
```{r}
umap <- umap::umap(pca_corrected$x[, 1:50])
```

```{r}


#  Extract UMAP coordinates
umap_coordinates <- as.data.frame(umap$layout)
colnames(umap_coordinates) <- c("UMAP1", "UMAP2")  

#  metadata for colouring like orig.ident or cell_type
umap_coordinates$cell_type <- fused_seurat$cell_type[rownames(umap_coordinates)]
umap_coordinates$orig.ident <- fused_seurat$orig.ident[rownames(umap_coordinates)]

#  Visualise with ggplot2
ggplot(umap_coordinates, aes(x = UMAP1, y = UMAP2, color = orig.ident)) +
  geom_point(alpha = 0.6) +  
  labs(title = "UMAP: Cell Type", x = "UMAP1", y = "UMAP2") +
  theme_minimal() 
  

```

# this is priority for now
use batch corrected
scale the data expression values by for each gene, use scale function (always by column . make genes by column)
calculate means scale data for each groupings of the cell types for all genes
for each of 4 groupings calculate mean expression for each
see what genes contribute to each

genes on rows, groupings as colums fro matrix then cluster




```{r}
# Check the row and column names
head(rownames(logcounts_matrix_rbe2))  # shows the first few row names

```

```{r}
head(colnames(logcounts_matrix_rbe2))  # shows the first few column names
```

# need to transpose as scale wants genes by columns

```{r}
# Scale the expression values genes for each gene across cells

scaled_logcounts_matrix_rbe2 <- scale(t(logcounts_matrix_rbe2))


#transposing it so genes become coulumns

```

```{r}
#head(t(logcounts_matrix_rbe2))[1:5, 1:5]
```

#checking the orientation
```{r}
head(colnames(scaled_logcounts_matrix_rbe2))
```

## need to calculate the means for each cell_type based from scaled_logcounts_matrix_rbe2
```{r}
the_cell_types <- fused_seurat$cell_type

```

```{r}

# Convert to data frame
scaled_df <- as.data.frame(scaled_logcounts_matrix_rbe2)

```

```{r}
head(scaled_df)
```


```{r}
# cell type as a new column
scaled_df$the_cell_type <- the_cell_types
```


```{r}
# calculating means for each gene within each cell type
means_by_cell_type <- scaled_df %>%
  group_by(the_cell_type) %>%
  summarise_all(mean, na.rm = TRUE)  #  mean for each gene across cells of the same type

```

```{r}
# Convert 
means_by_cell_type <- as.data.frame(means_by_cell_type)
```

```{r}
head(means_by_cell_type)
```



```{r}
#combined_mice_AIS_genes[! combined_mice_AIS_genes %in% c("Iqschfp", "Mfrp", "Foxa2", "Gyg1", "Nps")]

```

# getting the new list of AIS genes

```{r}
AIS_categories <- read.csv("/scratch/prj/dn_hamidlab_msc_proj/k24082190/originals/New_AIS_cat_prop.csv")
```


```{r}
tail(AIS_categories)
```



```{r}
AIS_categories[12, 5] <- "Enzyme"
```




```{r}
AIS_categories <- AIS_categories %>%
  filter(!grepl("look above", AIS_specific, ignore.case = TRUE))

```

```{r}
AIS_categories <- AIS_categories %>%
  filter(!grepl("look below", AIS_specific, ignore.case = TRUE))
```


```{r}
new_AIS_genes_df <- as.data.frame(AIS_categories)
```


```{r}
keeps <- c("Mouse_name", "AIS_specific")
```


```{r}
new_AIS_genes_df <- new_AIS_genes_df[keeps]
```




```{r}
head(new_AIS_genes_df)
```

```{r}
new_AIS_genes_df <- new_AIS_genes_df[new_AIS_genes_df$AIS_specific %in% c("yes", "maybe", "Yes", "Maybe"), ]

```


```{r}
new_AIS_genes_df[17, 1] <- "Kcnb2"

```



```{r}
new_AIS_genes <- as.character(new_AIS_genes_df$Mouse_name)
```

```{r}
new_AIS_genes
```



```{r}
filtered_means_by_cell_type <- means_by_cell_type[, c("the_cell_type", intersect(colnames(means_by_cell_type), new_AIS_genes))]

filtered_means_by_cell_type <- filtered_means_by_cell_type %>% column_to_rownames(var = "the_cell_type")

# check the resulting dataframe
head(filtered_means_by_cell_type)

```


# create heatmap
```{r}

filtered_means_by_cell_type_m <- as.matrix(filtered_means_by_cell_type)
heatmap(filtered_means_by_cell_type_m,  scale = "none")
```

```{r}

the_pheatmap <- pheatmap(filtered_means_by_cell_type_m,  scale = "none")
```

##### use cutree
save pheatmap to new variable then use $ to see slots to use for cutree, set k = 8
gene ontology for each grouping clusterProfiler, background genes is AIS, plot out scale expression of entire cluster in the 4 groupingd box plot or scatter plot of the genes in that grouping


```{r}
gene_clusters <- cutree(the_pheatmap$tree_col, k = 5)
```


```{r}


# Convert to a dataframe
cluster_df <- data.frame(Gene = names(gene_clusters), Cluster = gene_clusters)

# Check the first few rows
head(cluster_df)

```

#which genes correspond to the first group
```{r}
which(gene_clusters==1)
```


```{r}
which(gene_clusters==2)
```

```{r}
which(gene_clusters==3)
```

```{r}
which(gene_clusters==4)
```


```{r}
which(gene_clusters==5)
```




```{r}
ego2 <- enrichGO(gene         = cluster_df$Gene,
                universe      = names(new_AIS_genes),
                OrgDb         = org.Mm.eg.db,
                keyType       = 'SYMBOL',
                ont           = "All",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.01,
                qvalueCutoff  = 0.05)
head(ego2)    
```




#converting to entrez ids  (dont need to run)
```{r}
# Convert gene symbols to Entrez IDs
gene_conversion <- bitr(cluster_df$Gene, 
                        fromType = "SYMBOL", 
                        toType = "ENTREZID", 
                        OrgDb = org.Mm.eg.db)

# Check the converted IDs
head(gene_conversion)
```

#same here
```{r}
# Merge cluster info with Entrez IDs
cluster_entrez_df <- merge(cluster_df, gene_conversion, by.x = "Gene", by.y = "SYMBOL", all.x = TRUE)

# Check the merged dataframe
head(cluster_entrez_df)

```

#same here
```{r}
# Extract only the Entrez IDs
gene_entrez <- gene_conversion$ENTREZID
```




Cell types on x axis. Y-axis is scaled expression, back to the other dataframe
```{r}
#converting to long format because thats what ggplot2 wants
filtered_means_by_cell_type_ldf <- filtered_means_by_cell_type %>%
  rownames_to_column(var = "Cell_Type") %>%  # Move rownames into a column
  pivot_longer(cols = -Cell_Type, names_to = "Gene", values_to = "Scaled_Expression")

```

```{r}
head(filtered_means_by_cell_type_ldf)

```




```{r}
ggplot(filtered_means_by_cell_type_ldf, aes(x = Cell_Type, y = Scaled_Expression, fill = Cell_Type)) +
  geom_boxplot(alpha = 0.6) +
  labs(title = "Distribution of Scaled Expression by Cell Type",
       x = "Cell Type",
       y = "Scaled Expression") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```



```{r}
gene_clusters_ndf <- as.data.frame(gene_clusters)
head(gene_clusters_ndf)

clustered_split_genes <- split(rownames(gene_clusters_ndf), gene_clusters_ndf$gene_clusters)
```



```{r}


# Perform enrichGO for each cluster
ego_results_1 <- lapply(clustered_split_genes, function(genes) {
  enrichGO(
    gene         = genes,     
    universe = (new_AIS_genes),
    OrgDb        = org.Mm.eg.db,     
    keyType      = "SYMBOL",         
    ont          = "ALL",            
    pAdjustMethod= "BH",            
    pvalueCutoff = 0.5,             
    qvalueCutoff = 1              
  )
})

# results for the first cluster
ego_results_1[[1]]

```

```{r}
lapply(ego_results_1, head, n = 20)

```


```{r}
ggplot(filtered_means_by_cell_type_ldf, aes(x = Cell_Type, y = Scaled_Expression, fill = Cell_Type)) +
  geom_boxplot(alpha = 0.6) +
  labs(title = "Distribution of Scaled Expression by Cell Type",
       x = "Cell Type",
       y = "Scaled Expression") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#### to test out id working
```{r}
cluster_id <- 1  # 

ggplot(filtered_means_by_cell_type_ldf, aes(x = Cell_Type, y = Scaled_Expression, fill = Cell_Type)) +
  geom_boxplot(alpha = 0.6) +
  labs(title = paste("Distribution of Scaled Expression by Cell Type - Cluster", cluster_id),
       x = "Cell Type",
       y = "Scaled Expression") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```









```{r}


# filtering the main dataframe for each cluster
clustered_data <- lapply(clustered_split_genes, function(genes) {
  filtered_means_by_cell_type_ldf[filtered_means_by_cell_type_ldf$Gene %in% genes, ]
})

plot_list <- lapply(clustered_data, function(df) {
  ggplot(df, aes(x = Cell_Type, y = Scaled_Expression, fill = Cell_Type)) +
    geom_boxplot(alpha = 0.6) +
    labs(
      title = "Distribution of Scaled Expression by Cell Type",
      x = "Cell Type",
      y = "Scaled Expression"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
})

# View the first plot
print(plot_list[[1]])




```


### test out cluser id for all
```{r}
# filtering the main dataframe for each cluster
clustered_data <- lapply(clustered_split_genes, function(genes) {
  filtered_means_by_cell_type_ldf[filtered_means_by_cell_type_ldf$Gene %in% genes, ]
})

# Create plots with cluster numbers in the title
plot_list <- lapply(seq_along(clustered_data), function(i) {
  df <- clustered_data[[i]]
  
  ggplot(df, aes(x = Cell_Type, y = Scaled_Expression, fill = Cell_Type)) +
    geom_boxplot(alpha = 0.6) +
    labs(
      title = paste("Cluster", i),
      x = "",
      y = "Scaled Expression"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 11))+
    theme(legend.position = "none")
})

# View the first plot
print(plot_list[[1]])

```







```{r}
#  grid layout 
par(mfrow = c(2, 2))  

# plot each one
invisible(lapply(plot_list, print))

#reset layout
par(mfrow = c(1, 1))


## I had to use invisible function and reset the layout because r would produce a duplicate for each plot. 16 plots would be produced   

library(patchwork)
plot_list[[1]] + plot_list[[2]] + plot_list[[3]] + plot_list[[4]] + plot_list[[5]]
```

get the go terms and bar plot that
rvgo to get smaller proportion of terms

https://stackoverflow.com/questions/43100566/using-seq-along-and-lapply-to-process-multiple-dataframes-capm 
#come back to this
```{r}


# Use lapply to loop through each cluster and plot
lapply(seq_along(ego_results_1), function(i) {
  barplot(
    ego_results_1[[i]], 
    showCategory = 8,     # Show top 10 categories
    title = paste("Cluster", i, "Enriched GO Terms")
  )
})

```


```{r}
lapply(seq_along(ego_results_1), function(i) {
  mutate(ego_results_1[[i]], qscore = -log(p.adjust, base=10)) %>% 
    barplot(x="qscore")})
```


```{r}
lapply(seq_along(ego_results_1), function(i) {
  mutate(ego_results_1[[i]], pvalue = pvalue) %>% 
    barplot(x="pvalue")})
```



```{r}


lapply(seq_along(ego_results_1), function(i) {
  df <- as.data.frame(ego_results_1[[i]])

  # Get top 10 GO terms by FoldEnrichment
  top_df <- df %>%
    slice_max(order_by = FoldEnrichment, n = 10)

  ggplot(top_df, aes(x = reorder(Description, FoldEnrichment),
                     y = FoldEnrichment,
                     fill = FoldEnrichment)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    scale_fill_gradient(low = "lightblue", high = "red") +
    labs(
      title = paste("Cluster", i, "Enriched GO Terms"),
      x = "GO Term",
      y = "Fold Enrichment"
    ) +
    theme_minimal()
})

```


```{r}


lapply(seq_along(ego_results_1), function(i) {
  df <- as.data.frame(ego_results_1[[i]])

  # Get top 10 GO terms by FoldEnrichment
  top_df <- df %>%
    slice_max(order_by = FoldEnrichment, n = 10)

  ggplot(top_df, aes(x = reorder(Description, FoldEnrichment),
                     y = FoldEnrichment,
                     fill = FoldEnrichment)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    scale_fill_gradient(low = "lightblue", high = "red") +
    labs(
      title = paste("Cluster", i, "Enriched GO Terms"),
      x = "GO Term",
      y = "Fold Enrichment"
    ) +
    theme_minimal()+
    theme(axis.text=element_text(size=11),
        axis.title=element_text(size=13))
})

```

```{r}
#library(stringr)

lapply(seq_along(ego_results_1), function(i) {
  df <- as.data.frame(ego_results_1[[i]])

  # Clean up GO term labels
  top_df <- df %>%
    mutate(
      Description = Description %>%
        str_replace(" activity$", "") %>%
        str_replace_all("transporter", "trans") %>%
        str_replace_all("channel", "chan")
    ) %>%
    slice_max(order_by = FoldEnrichment, n = 10)

  ggplot(top_df, aes(x = reorder(Description, FoldEnrichment),
                     y = FoldEnrichment,
                     fill = FoldEnrichment)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    scale_fill_gradient(low = "lightblue", high = "red") +
    labs(
      title = paste("Cluster", i, "Enriched GO Terms"),
      x = "GO Term",
      y = "Fold Enrichment"
    ) +
    theme_minimal() +
    theme(
      axis.text.y = element_text(size = 12),
      axis.text.x = element_text(size = 12)
    )
})

```


```{r}


# Create list of GO barplots
go_plots <- lapply(seq_along(ego_results_1), function(i) {
  df <- as.data.frame(ego_results_1[[i]])

  # Clean up GO term labels
  top_df <- df %>%
    mutate(
      Description = Description %>%
        str_replace(" activity$", "") %>%
        str_replace_all("transporter", "trans") %>%
        str_replace_all("channel", "chan")
    ) %>%
    slice_max(order_by = FoldEnrichment, n = 10)

  ggplot(top_df, aes(x = reorder(Description, FoldEnrichment),
                     y = FoldEnrichment,
                     fill = FoldEnrichment)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    scale_fill_gradient(low = "lightblue", high = "red") +
    labs(
      title = paste("Cluster", i, "Enriched GO Terms"),
      x = "GO Term",
      y = "Fold Enrichment"
    ) +
    theme_minimal() +
    theme(
      axis.text.y = element_text(size = 12),
      axis.text.x = element_text(size = 12)
    )
})

# Combine the first two plots using patchwork
go_plots[[1]] / go_plots[[2]]

```



















### now going to use the csv file that has the list of genes sorted into categories.

```{r}
AIS_categories <- read.csv("/scratch/prj/dn_hamidlab_msc_proj/k24082190/originals/New_AIS_cat_prop.csv")
```


```{r}
tail(AIS_categories)
```


```{r}
AIS_categories[12, 5] <- "Enzyme"
```

```{r}
AIS_categories <- AIS_categories %>%
  filter(!grepl("look above", AIS_specific, ignore.case = TRUE))

```

```{r}
AIS_categories <- AIS_categories %>%
  filter(!grepl("look below", AIS_specific, ignore.case = TRUE))
```


```{r}
tail(AIS_categories)
```


```{r}
AIS_categories_df <- as.data.frame(AIS_categories)
```



```{r}
AIS_categories_df <- AIS_categories_df[, -1]

```


```{r}
AIS_categories_df[17, 1] <- "Kcnb2"

```


```{r}


# result list to store summaries for each cluster
cluster_summary <- lapply(clustered_split_genes, function(cluster_genes) {
  
  # summary data frame for each cluster
  summary <- as.data.frame(sapply(unique(AIS_categories_df$Category), function(cat) {
    
    # Get all genes in the current category
    genes_in_cat <- AIS_categories_df$Mouse_name[AIS_categories_df$Category == cat]
    
    # Count how many genes in the cluster are part of this category
    part_of <- sum(genes_in_cat %in% cluster_genes)
    
    # Count how many genes in the cluster are not part of this category
    genes_not_in_cat <- AIS_categories_df$Mouse_name[AIS_categories_df$Category != cat]
    not_part_of <- sum(genes_not_in_cat %in% cluster_genes)
    
    # Return both counts: Part of and Not part of for this category
    return(c("Part of" = part_of, "Not part of" = not_part_of))
  }))
  
  # Return the summary for the current cluster
  return(summary)
})



```

```{r}
# View all cluster summaries
cluster_summary

```


```{r}
## same as above but having background not background columns

# result list to store summaries for each cluster
cluster_summary_2 <- lapply(clustered_split_genes, function(cluster_genes) {
  
  # Create a summary data frame for each cluster
  summary <- as.data.frame(sapply(unique(AIS_categories_df$Category), function(cat) {
    
    # Genes in this category
    genes_in_cat <- AIS_categories_df$Mouse_name[AIS_categories_df$Category == cat]
    genes_not_in_cat <- AIS_categories_df$Mouse_name[AIS_categories_df$Category != cat]
    
    # Cluster-level counts
    part_of <- sum(cluster_genes %in% genes_in_cat)
    not_part_of <- sum(cluster_genes %in% genes_not_in_cat)
    
    # Background counts
    background_total <- length(unique(genes_in_cat))
    not_background_total <- length(unique(genes_not_in_cat))
    
    return(c(
      "Part of" = part_of,
      "Not part of" = not_part_of,
      "Background" = background_total,
      "Not background" = not_background_total
    ))
  }))
  
  return(summary)
})

```


```{r}
cluster_summary_2
```

```{r}
cluster_summary_2_t <- lapply(clustered_split_genes, function(cluster_genes) {
  categories <- unique(AIS_categories_df$Category)
  
  summary <- sapply(categories, function(cat) {
    genes_in_cat <- AIS_categories_df$Mouse_name[AIS_categories_df$Category == cat]
    genes_not_in_cat <- AIS_categories_df$Mouse_name[AIS_categories_df$Category != cat]
    
    part_of <- sum(cluster_genes %in% genes_in_cat)
    not_part_of <- sum(cluster_genes %in% genes_not_in_cat)
    
    background_total <- length(unique(genes_in_cat))
    not_background_total <- length(unique(genes_not_in_cat))
    
    return(c(part_of, not_part_of, background_total, not_background_total))
  })
  
  # Transpose and format properly
  summary_df <- as.data.frame(t(summary), stringsAsFactors = FALSE)
  colnames(summary_df) <- c("Part of", "Not part of", "Background", "Not background")
  summary_df$Category <- rownames(summary_df)
  rownames(summary_df) <- NULL
  
  # Move 'Category' to the front
  summary_df <- summary_df[, c("Category", "Part of", "Not part of", "Background", "Not background")]
  
  # Make sure all numeric columns are numeric
  summary_df[ , 2:5] <- lapply(summary_df[ , 2:5], as.numeric)
  
  return(summary_df)
})




```

```{r}
cluster_summary_2_t
```


## running the fishers test
```{r}
Fishers_test <- lapply(cluster_summary_2_t, function(df){
  out <- bind_rows(apply(df,1, function(line){
    dat <- matrix(line, nrow = 2, byrow = T)
    fisher.out <- fisher.test(dat)
    data.frame(pvalue = fisher.out$p.value,
               estimate = fisher.out$estimate)
  }))
  rownames(out) <- rownames(df)
  out
})
```

fisher.test() is being given a matrix (dat) that includes invalid values e.g. NA, negative, or non-numeric values.

line is a row of a data.frame, and it includes non-numeric data, like the Category column which is a string datatype
need to only use the numeric columns when building matrix


```{r}
Fishers_test2 <- lapply(cluster_summary_2_t, function(df) {
  out <- bind_rows(apply(df, 1, function(line) {
    
    # Convert the values to numeric 
    part_of <- as.numeric(line["Part of"])
    not_part_of <- as.numeric(line["Not part of"])
    background <- as.numeric(line["Background"])
    not_background <- as.numeric(line["Not background"])
    
    #  2x2 contingency table
    dat <- matrix(c(part_of,
                    not_part_of,
                    background - part_of,
                    not_background - not_part_of),
                  nrow = 2, byrow = TRUE)
    
    # Fisher test
    fisher.out <- fisher.test(dat)
    
    # Return p-value and odds ratio
    data.frame(pvalue = fisher.out$p.value,
               estimate = fisher.out$estimate)
  }))
  
  rownames(out) <- df$Category
  out
})


```


```{r}
Fishers_test2[[1]]  # Cluster 1 etc
Fishers_test2[[2]] 
Fishers_test2[[3]]  
Fishers_test2[[4]]  
Fishers_test2[[5]]  

```

#Merging the above dataframes for the clusters, adding identitiies to them and then making a dot plot where categories is on the y-axis, x axis is the clusters, the size of the dot correlate to the estimate value and colour correlates to the p-value.


```{r}
# Add cluster identities to each data frame
Fishers_test2_with_cluster <- lapply(seq_along(Fishers_test2), function(i) {
  df <- Fishers_test2[[i]]
  df$Cluster <- paste("Cluster", i)
  return(df)
})

```

```{r}
#merging the dataframes
combined_fishers_df <- do.call(rbind, lapply(Fishers_test2_with_cluster, rownames_to_column, "Category" ))
```


```{r}
# Convert the 'Category' and 'Cluster' columns to factors for plotting as ggplot2, often treats categorical data as factors

combined_fishers_df$Category <- factor(combined_fishers_df$Category, 
                                    levels = unique(combined_fishers_df$Category))
combined_fishers_df$Cluster <- factor(combined_fishers_df$Cluster, 
                                   levels = paste("Cluster", 1:5))
```

# change to -log10 of pvalue

```{r}
ggplot(combined_fishers_df, aes(x = Cluster, y = Category, size = estimate, color = pvalue)) +
  geom_point()+
  scale_size_continuous(range = c(3, 10)) +  #dot size
  scale_color_gradient(low = "blue", high = "red") + #colour gradient
  theme_minimal() +
  labs(
    title = "Fisher's Test results, clusters against categories",
    x = "Cluster",
    y = "Category",
    size = "Estimate",
    color = "pvalue"
    
  )
  
```

#Fisher test results
```{r}
ggplot(combined_fishers_df, aes(x = Cluster, y = Category, size = estimate, color = -log10(pvalue))) +
  geom_point() +
  scale_size_continuous(range = c(3, 10)) +  # Dot size
  scale_color_gradient(low = "blue", high = "red") +  # High = more significant
  theme_minimal() +
  labs(
    title = "Cluster Enriched Categories",
    x = "Cluster",
    y = "Category",
    size = "Estimate",
    color = "-log10(p-value)"
  )

```





###Pseudotime   #################################################

```{r}
obj_harmony <- readRDS("/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/obj_harmony.rds")
```



```{r}
data <- Embeddings(obj_harmony, "umap.harmony")
dm <- DiffusionMap(data, n_pcs = 3)

```

```{r}
mapp_diff <- obj_harmony[["diffmap"]] <- CreateDimReducObject(embeddings = dm@eigenvectors[,1:3], key = "dc_", assay = "SCT")
DimPlot(obj_harmony, reduction = "diffmap", label = TRUE) 
```


```{r}
sds <- slingshot(Embeddings(obj_harmony, "diffmap"), 
                 clusterLabels = obj_harmony$seurat_clusters,
                 start.clus= '13')
```



```{r}
slingLineages(sds)
```

```{r}
obj_harmony$pseudotime <- slingAvgPseudotime(sds)
```

```{r}
FeaturePlot(obj_harmony, "pseudotime", 
            reduction = "diffmap", label = T)
```

```{r}
 FeaturePlot(obj_harmony, ("pseudotime"), 
            reduction = "umap.harmony", label = T)
```


```{r}
FeaturePlot(obj_harmony, c("TopOSN_Score1", "TopOSN_Score2"), 
            reduction = "umap.harmony", label = T)
```

```{r}
plots <- FeaturePlot(obj_harmony, features = c("TopOSN_Score1", "TopOSN_Score2"), 
            reduction = "diffmap", label = T, combine = F) 
lapply(plots, function(plot)  {
  plot + 
  scale_x_continuous(limits = c(-0.025,0.025)) +
     scale_y_continuous(limits = c(-0.025,0.025))
})
```

```{r}
# Create FeaturePlot objects without combining
plots <- FeaturePlot(obj_harmony, 
                     features = c("TopOSN_Score1", "TopOSN_Score2"), 
                     reduction = "diffmap", 
                     label = TRUE, 
                     combine = FALSE)

# Adjust limits and combine with patchwork
adjusted_plots <- lapply(plots, function(plot) {
  plot +
    scale_x_continuous(limits = c(-0.025, 0.025)) +
    scale_y_continuous(limits = c(-0.025, 0.025))
})

# Arrange side by side
wrap_plots(adjusted_plots, ncol = 1)
```











```{r}
plot <- FeaturePlot(obj_harmony, features = ("pseudotime"), 
            reduction = "diffmap", label = T, combine = F) 
lapply(plots, function(plot)  {
  plot + 
  scale_x_continuous(limits = c(-0.025,0.025)) +
     scale_y_continuous(limits = c(-0.025,0.025))
})
```

```{r}
# Recreate the proper plots
diffmap_plot <- DimPlot(obj_harmony, reduction = "diffmap", label = TRUE)

# Feature plots for TopOSN scores 
plots <- FeaturePlot(obj_harmony, features = c("TopOSN_Score1", "TopOSN_Score2"), 
                     reduction = "diffmap", label = TRUE, combine = FALSE)
plots <- lapply(plots, function(plot) {
  plot + 
    scale_x_continuous(limits = c(-0.025,0.025)) +
    scale_y_continuous(limits = c(-0.025,0.025))
})

# Pseudotime FeaturePlot
plot_pseudo <- FeaturePlot(obj_harmony, features = "pseudotime", 
                           reduction = "umap.harmony", label = TRUE)

# Combine everything properly
final_plot <- diffmap_plot + plots[[1]] + plots[[2]]
final_plot
```






# reloading this rds to get the annotations of which are immature and which are mature
```{r}
fused_seurat <- readRDS("/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/fused_seurat.rds")
```


## goint to annotate the obj_harmony to see which are immature and which are mature
```{r}
##extracting the cell type annotations from fused_seurat

neuron_labels <- fused_seurat@meta.data$cell_type
```


```{r}
#naming the vector with cell barcodes
names(neuron_labels) <- colnames(fused_seurat)
```


```{r}
#adding cell type annotation to obj_harmony
obj_harmony$neuron_type <- neuron_labels[colnames(obj_harmony)]
```


## plotting density plot for pseudotime
```{r}
df_dp <- obj_harmony@meta.data[, c("pseudotime", "neuron_type")]
#df_dp <- df[!is.na(df$neuron_type), ] #removing cells without a label
```

```{r}
ggplot(df_dp, aes(x = pseudotime, color = neuron_type, fill = neuron_type)) +
  geom_density(alpha = 0.3) +
  labs(
    x = "Pseudotime",
    y = "Density",
    color = "Neuron Type",
    fill = "Neuron Type"
  ) +
  ggtitle("Density Plot of Pseudotime") +
  theme_minimal()

```


#using geom_histogram
```{r}
ggplot(df_dp, aes(x = pseudotime, fill = neuron_type))+
  geom_histogram(alpha = 0.5, position = "identity", bins = 20)+
  labs(x = "pseudotime", y = "", color = "neuron_type", fill = "neuron_type")+
  theme_minimal()
  
```

# add an new column in the metadata for the pseudorank
```{r}
obj_harmony$pseudorank <- rank(obj_harmony$pseudotime, ties.method = "average")
```


#normalising the pseudorank column
```{r}
n_cells <- ncol(obj_harmony)
obj_harmony$pseudorank_scaled <- (obj_harmony$pseudorank * 10) / n_cells

```


```{r}
df_s <- obj_harmony@meta.data[, c("pseudorank_scaled", "neuron_type")]
```



```{r}
ggplot(df_s, aes(x = pseudorank_scaled, color = neuron_type, fill = neuron_type))+
  geom_density(alpha = 0.3)+
  labs(x = "pseudorank", y = "density", color = "neuron_type", fill = "neuron_type")+
  theme_minimal()
```





```{r}
FeatureScatter(obj_harmony, "pseudotime", "Nav1") +
  geom_smooth()
```

```{r}
weights <- slingCurveWeights(sds)
counts <- GetAssayData(obj_harmony, assay = "SCT", slot = "counts")
```

#going to make weights have one column by taking the highest value in across the 4 lineages
```{r}
weights <- rowMaxs(weights)
```


# back to a dataframe
```{r}
weights_df <- as.data.frame(weights)
```


## sampling random 5000 cells from obj_harmony to run the fitgam function on
```{r}
set.seed(321)

sampled_cells <- sample(colnames(obj_harmony), 5000)

obj_harmony_subset <- subset(obj_harmony, cells = sampled_cells)

counts <- obj_harmony_subset@assays$SCT@counts
pseudotime <- obj_harmony_subset[["pseudotime"]]
cellWeights <- weights[sampled_cells]

sce <- fitGAM(counts = counts, 
              pseudotime = pseudotime, 
              cellWeights = cellWeights,
              )

```


#saving rds file so i dont have to run fitgam again
```{r}
#saveRDS(sce, file = "/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/sce.rds")
sce <- readRDS("/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/sce.rds")
```





#skip
```{r}
FeaturePlot(obj_harmony_subset, ("pseudotime"), 
            reduction = "umap", label = T)
```
#skip
```{r}
FeaturePlot(obj_harmony_subset, c("TopOSN_Score1", "TopOSN_Score2"), 
            reduction = "umap.harmony", label = T)
```
#skip
```{r}
plots <- FeaturePlot(obj_harmony_subset, features = c("TopOSN_Score1", "TopOSN_Score2"), 
            reduction = "diffmap", label = T, combine = F) 
lapply(plots, function(plot)  {
  plot + 
  scale_x_continuous(limits = c(-0.025,0.025)) +
     scale_y_continuous(limits = c(-0.025,0.025))
})
```
#skip
```{r}
plots <- FeaturePlot(obj_harmony_subset, features = ("pseudotime"), 
            reduction = "diffmap", label = T, combine = F) 
lapply(plots, function(plot)  {
  plot + 
  scale_x_continuous(limits = c(-0.025,0.025)) +
     scale_y_continuous(limits = c(-0.025,0.025))
})
```
#skip
#startVsEndTest uses a Wald test to check if the average gene activity at the beginning equals that at the end
```{r}
startRes <- startVsEndTest(sce)
```

#skip
#top and bottom 10 differentially expressed genes that are statistically significant and with high Wald’s statistics
```{r}
startRes %>% 
  filter(pvalue < 0.05) %>% 
  filter(waldStat > 100) %>% 
  slice_max(logFClineage1, n = 10)
```

#skip
# 10 bottom most
```{r}
startRes %>% 
  filter(pvalue < 0.05) %>% 
  filter(waldStat > 100) %>% 
  slice_min(logFClineage1, n = 10)
```


```{r}
FeatureScatter(obj_harmony_subset, "pseudotime", "Prxl2b") +
    geom_smooth()
```

```{r}
FeatureScatter(obj_harmony_subset, "pseudotime", "Mfap2") +
    geom_smooth()
```

#skip
```{r}
assoRes <- associationTest(sce)
head(assoRes)
```

## load in rds file created from local
```{r}
start_asso_df <- readRDS("/scratch/prj/dn_hamidlab_msc_proj/k24082190/Outputs/start_asso_df.rds")
head(start_asso_df)
```

```{r}
arrange(start_asso_df, desc(waldStat))
```

#trying to find those that peak around 0.04

Eif1, Cep250, Tulp4, Ncoa1
```{r}
FeatureScatter(obj_harmony_subset, "pseudotime", "Ncoa1") + geom_smooth()
```



```{r}
p1 <- FeatureScatter(obj_harmony_subset, "pseudotime", "Ncoa1") + 
  geom_smooth() + 
  theme(legend.position = "none")

p2 <- FeatureScatter(obj_harmony_subset, "pseudotime", "Tulp4") + 
  geom_smooth() + 
  theme(legend.position = "none")

p3 <- FeatureScatter(obj_harmony_subset, "pseudotime", "Cep250") + 
  geom_smooth() + 
  theme(legend.position = "none")

p4 <- FeatureScatter(obj_harmony_subset, "pseudotime", "Eif1") + 
  geom_smooth() + 
  theme(legend.position = "none")

# Combine and display
p1 + p2 + p3 + p4

```





## creating plots
```{r}
# selected gene is Mgst3      #change this for any gene want to plot
selected_gene <- "Mgst3"

# taking pseudotime from metadata
pseudotime <- obj_harmony_subset@meta.data$pseudotime

# Pull gene expression
gene_expression <- FetchData(obj_harmony_subset, vars = selected_gene)[[1]]

# Combine into a data frame
plot_data <- data.frame(
  pseudotime = pseudotime,
  expression = gene_expression
)
  
```


```{r}
ggplot(plot_data, aes(x = pseudotime, y = expression)) +
  geom_point(alpha = 0.5) +  
  geom_smooth(method = "loess", color = "blue") +  
  labs(
    title = paste("Expression of", selected_gene, "over Pseudotime"),
    x = "Pseudotime",
    y = "Gene Expression"
  ) +
  theme_minimal()
```

Nenf, Gng3, Cst3, Syt5
```{r}
FeatureScatter(obj_harmony_subset, "pseudotime", "Nenf") + geom_smooth()
```


## get rid of legends plot_layout
```{r}
p1 <- FeatureScatter(obj_harmony_subset, "pseudotime", "Nenf") + geom_smooth()
p2 <- FeatureScatter(obj_harmony_subset, "pseudotime", "Gng3") + geom_smooth()
p3 <- FeatureScatter(obj_harmony_subset, "pseudotime", "Cst3") + geom_smooth()
p4 <- FeatureScatter(obj_harmony_subset, "pseudotime", "Syt5") + geom_smooth()      

# Combine and display all
p1 + p2 + p3 + p4             ##using patchwork 
```


```{r}
p1 <- FeatureScatter(obj_harmony_subset, "pseudotime", "Nenf") + 
  geom_smooth() + 
  theme(legend.position = "none")

p2 <- FeatureScatter(obj_harmony_subset, "pseudotime", "Gng3") + 
  geom_smooth() + 
  theme(legend.position = "none")

p3 <- FeatureScatter(obj_harmony_subset, "pseudotime", "Cst3") + 
  geom_smooth() + 
  theme(legend.position = "none")

p4 <- FeatureScatter(obj_harmony_subset, "pseudotime", "Syt5") + 
  geom_smooth() + 
  theme(legend.position = "none")

# Combine and display
p1 + p2 + p3 + p4

```



```{r}
new_AIS_genes
```




Ank3, Nav1, Ank2, Fgf12
```{r}
FeatureScatter(obj_harmony_subset, "pseudotime", "Fgf12") + geom_smooth()
```


```{r}

p4 <- FeatureScatter(obj_harmony_subset, "pseudotime", "Ank3") + 
  geom_smooth() + 
  theme(legend.position = "none")

p5 <- FeatureScatter(obj_harmony_subset, "pseudotime", "Nav1") + 
  geom_smooth() + 
  theme(legend.position = "none")

p6 <- FeatureScatter(obj_harmony_subset, "pseudotime", "Ank2") + 
  geom_smooth() + 
  theme(legend.position = "none")

p7 <- FeatureScatter(obj_harmony_subset, "pseudotime", "Fgf12") + 
  geom_smooth() + 
  theme(legend.position = "none")

# Combine and display
p4 + p5 + p6 +p7

```
```








```{r}
p1 <- FeatureScatter(obj_harmony_subset, "pseudotime", "Nav2") + 
  geom_smooth() + 
  theme(legend.position = "none")

p2 <- FeatureScatter(obj_harmony_subset, "pseudotime", "Map2") + 
  geom_smooth() + 
  theme(legend.position = "none")

p3 <- FeatureScatter(obj_harmony_subset, "pseudotime", "Trim46") + 
  geom_smooth() + 
  theme(legend.position = "none")

#p4 <- FeatureScatter(obj_harmony_subset, "pseudotime", "Nrcam") + 
  #geom_smooth() + 
  #theme(legend.position = "none")

p5 <- FeatureScatter(obj_harmony_subset, "pseudotime", "Kcnb2") + 
  geom_smooth() + 
  theme(legend.position = "none")

p6 <- FeatureScatter(obj_harmony_subset, "pseudotime", "L1cam") + 
  geom_smooth() + 
  theme(legend.position = "none")

p7 <- FeatureScatter(obj_harmony_subset, "pseudotime", "Scn3a") + 
  geom_smooth() + 
  theme(legend.position = "none")

# Combine and display
p1 + p2 + p3 + p5 + p6 +p7

```































